---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

<style>
.nobr {
  white-space: nowrap;
}
</style>


```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "polishing")
```

## Comments on data by animal species

### Pigs {#sec-sales-pigs}

Antibiotics for pigs are predominantly sold on veterinary prescription by pharmacies to the animal owner and the intended target species is recorded by the dispensing pharmacy. Sales reported by pharmacies as prescriptions for pigs are therefore believed to closely reflect sales for commercial herds. 

In 2016 and 2025, the sales of antibiotics for pigs (pet pigs included) were <span class="nobr">2 874</span> and <span class="nobr">3 297</span> kg active substance, respectively, or 12.3 and 13.2 mg/kg slaughtered pig. Around 76% of the total sales for pigs were products for use in individual animals, and of these, 67% were products containing benzylpenicillin. Sales of fluoroquinolones were negligible, and no cephalosporins were sold for pigs in 2025. In Sweden, products with polymyxins (colistin) for oral use are only used for pigs. As noted under @sec-sales-population Population corrected data, a marked decrease in sales has been noted since 2016 and sales are today very low and most likely limited to cases where no alternatives are available. 

The apparent increase in total consumption is largely explained by an increased need to treat post-weaning diarrhoea following the 2022 withdrawal of products with high levels of zinc oxide, which were previously used in feed for diarrhoea prevention. Sales of products containing aminoglycosides for oral group treatment have increased, and notably, the occurrence of neomycin resistance in *Escherichia coli* from pigs (see [Antibiotic resistance in animals, Clinical isolates from Animals, Pigs](#sec-clinical-pigs) has recently necessitated the introduction of apramycin use in Sweden. 

The sales of relevant products, containing neomycin, paromomycin, apramycin or colistin for oral group treatment, classified under ATCvet code QA07AA and all sold on special license, both on prescription requisition, have been explored. Course doses were calculated by multiplying the dosage indicated in the product information of each product by an assumed standard weight of 12 kg for weaner pigs and a treatment duration of 3 days. The number of course doses sold in one month was divided by the number of pigs slaughtered five months later and the average number of course doses per 1 000 slaughtered pigs per quarter from 2019 to Q1 2026 is shown in @fig-23. When interpreting the data, it is important to note that actual use of the products may take place later than the sales.  


```{r}
#| include: false
# hidden chunk to read captions
dat_23 <- load_from_excel("data/Figure 2.3 Quarterly sales.xlsx", sheet = "Data")
fig_23_cap <- dat_23$fig_cap
fig_23_alt <- dat_23$fig_alt 

```

::: panel-tabset
###### Figure

```{r}
#| label: fig-23
#| message: false
#| warning: false
#| fig-cap: !expr fig_23_cap
#| fig-alt: !expr fig_23_alt

dat_23_inter <- dat_23$fig_data |>
  dplyr::select(-Year) |>
  pivot_longer(cols = c(`Prescribed course doses`, `Requisitioned course doses`), names_to = "Treatment type", values_to = "Value") 


dat_23_long <- dat_23_inter |>  
  mutate(stack_order = factor(`Treatment type`, levels = c("Requisitioned course doses", "Prescribed course doses"))) |>
  mutate(tooltip = ifelse(is.na(Value), paste0("<b>Year and quarter:</b> ", Quarter, "<br>No data"), 
                          paste0("<b>Year and quarter:</b> ", Quarter, "<br><b>Type:</b> ", `Treatment type`, "<br><b>Value:</b> ", round(Value, 1))), 
         data_id = paste(Quarter, `Treatment type`, sep = "_"))


p_23 <- ggplot(dat_23_long, 
               aes(x = Quarter, 
                   y = Value,
                   fill = `Treatment type`, 
                   group = stack_order,
                   data_id = data_id)) +
  ggiraph::geom_bar_interactive(aes(tooltip = tooltip), 
                                stat = "identity", position = "stack") +
  scale_x_discrete(labels = \(x) gsub(" Q", "\nQ", x)) +
  scale_fill_SweSva() +
  labs(x = NULL,
       y = "Average number of course doses/1000 slaughtered pigs", 
       fill = NULL) +
  scale_y_continuous(limits = c(0, 250), breaks = seq(0, 250, by = 50), expand=c(0,0.2)) +
  theme_SweSva() +
  theme(axis.text.x = element_text(size = 9, hjust = 0.5, vjust = 0.5, margin = margin(t = 5))) +
  add_sva_caption()

make_dynamic_plot(plot = p_23, filename = "Fig2-3", width = 10)

```

###### Data

```{r}
dat_23_inter |>
  make_table_data_tab(file_name = "Fig2-3")
```
:::


### Chickens

Antibiotics are rarely used for treatment of bacterial diseases in commercially reared *Gallus gallus*. Localised outbreaks can therefore have a major influence on sales in a specific year. Over the last ten years, the yearly sales of fluoroquinolones for slaughter chickens and hens have been below, or well below, 0.25 kg. Cephalosporins or colistin have never been used in broiler production in Sweden. 

From 2011, the Swedish poultry meat association requests all treatments of broilers, parents and grandparents to be reported as part of the Poultry health control programme. The programme covers more than 98% of the broilers reared in commercial production. The number of treated flocks and the number of flocks produced per year are shown in @tbl-22. In 2025, 30 flocks were reported as treated, which is considerably higher compared to the previous five years. Of these, 21 flocks were administered amoxicillin against colibacillosis, eight flocks were administered trimethoprim-sulphonamide against colibacillosis and one flock was administered phenoxymethylpenicillin against necrotic enteritis. In addition, seventeen parent flocks were treated, mainly against arthritis; 15 flocks were administered phenoxymethylpenicillin, one was administered trimethoprim-sulphonamide and one was administered amoxicillin. 

::: {.panel-tabset}
###### Table
```{r}
#| label: tbl-22
#| echo: false
#| message: false
#| warning: false

# Read source data from the Excel sheet
src_22 <- load_from_excel("data/Table 2.2 Chicken flocks.xlsx", sheet = "Data")

tab_22 <- make_qt_table_from_excel("data/Table 2.2 Chicken flocks.xlsx", sheet = "Data")
tab_22$table

```


###### Data
```{r}
df <- src_22$fig_data
df |>
  make_table_data_tab(file_name = "Table2-2")
```
:::

The reported antibiotic use in 2025 corresponds to 0.08 mg active substance per kg slaughtered chicken. The most common substance used was trimethoprim-sulphonamide. 

Coccidiostats of the ionophore group are used as feed additives to control coccidiosis in the production of chickens for slaughter and in turkeys. Since the late 1980s, narasin has been by far the most widely used substance for broilers. 

### Fish {#sec-sales-fish}

Medicated feed for fish is always traded from other Nordic countries. Therefore, the quantities sold are not captured by the national pharmacy sales collected by the Swedish eHealth Agency. Records of prescription of veterinary medicines for fish are collected annually by the veterinarian coordinating the limited number of veterinarians that deal with farmed fish, and the results are reported annually to the Swedish Board of Agriculture. However, it is not known whether the full amount prescribed was actually dispensed.

The occurrence of bacterial disease in farmed fish is influenced by water temperatures in summer and the amounts prescribed may therefore vary between the years. Antibiotics prescribed in 2025 were products with florfenicol, oxolinic acid, oxytetracycline and trimethoprim-sulphonamide. Florfenicol is primarily used for the treatment of flavobacteriosis (caused by *Flavobacterium psychrophilum*), a disease mainly affecting juveniles, i.e. fish of very low weight. Oxolinic acid, oxytetracycline and trimethoprim-sulphonamide are used to treat diseases caused by *Aeromonas salmonicida* and *Flavobacterium columnare*. These are diseases affecting production fish, i.e. of a higher weight. Therefore, the relations between the antibiotics shown in @fig-24 do not translate to treatment frequencies or actual exposure of individual fish. The dosage differs considerably between substances. Therefore, the prescribed amounts of active substance have been recalculated into quantities of medicated feed based on the commonly prescribed concentration. In @fig-24, the prescription of antibiotics for fish is shown as kg medicated feed per tonne of fish produced.  


```{r}
#| include: false
# hidden chunk to read captions
dat_24 <- load_from_excel("data/Figure 2.4 Antibiotics for fish.xlsx", sheet = "Data")
fig_24_cap <- dat_24$fig_cap
fig_24_alt <- dat_24$fig_alt 
```

::: panel-tabset
###### Figure

```{r}
#| label: fig-24
#| message: false
#| warning: false
#| fig-cap: !expr fig_24_cap
#| fig-alt: !expr fig_24_alt

dat_24_long <- dat_24$fig_data |> 
  pivot_longer(cols = -Year, names_to = "Antibiotic", values_to = "Value", values_drop_na = TRUE) |>
  mutate(Antibiotic = factor(Antibiotic, levels = c("Florfenicol", "Oxolinic acid", "Oxytetracycline", "Trimethoprim-sulphonamide")))


df_fig_24 <- dat_24_long |>
  mutate(stack_order = dplyr::case_when(Antibiotic == "Florfenicol" ~ 4,
                                        Antibiotic == "Oxolinic acid" ~ 3,
                                        Antibiotic == "Oxytetracycline" ~ 2,
                                        Antibiotic == "Trimethoprim-sulphonamide" ~ 1),
         tooltip = paste0("<b>Year:</b> ", Year,
                          "<br><b>Antibiotic:</b> ", `Antibiotic`, 
                          "<br><b>Value:</b> ", round(Value, 2), "Kg medicated feed/tonne of liveweight fish slaughtered"), 
         data_id = paste(`Antibiotic`, Year, sep = "_"))

p_24 <- ggplot(df_fig_24,
               aes(x = factor(Year),
                   y = Value, 
                   fill = `Antibiotic`, 
                   group = stack_order,
                   data_id = data_id)) +
  geom_bar_interactive(aes(tooltip = tooltip), stat = "identity", position = "stack") +
  scale_fill_SweSva() +
  scale_y_continuous(limits = c(0, 9), 
                     breaks = seq(0, 10, by = 1), 
                     expand=c(0,0.2)) +
  scale_x_discrete(expand=c(0,0.2)) +
  labs(x = "Year",
       y = "Kg medicated feed/tonne of liveweight fish slaughtered", fill = NULL) +
  theme_SweSva() +
  #theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  add_sva_caption()
make_dynamic_plot(plot = p_24, filename = "Fig2-4")

```

###### Data

```{r}
dat_24_long |>
  make_table_data_tab(file_name = "Fig2-4")
```
:::


### Horses

In 2025, sales of trimethoprim-sulphonamides formulated for oral use in horses (paste or powder) were 11% of the total sales of antibiotics for animals and 67% of the sales of all products with trimethoprim-sulphonamides. Since 2016, there has been a decrease in sales of trimethoprim-sulphonamides formulated for oral use in horses by 6%, measured in kg active substance. It is unclear if this decrease reflects increased adherence to guidelines or simply a decreasing number of horses, given that the latest population estimate is from 2016. 

The sales of other antibiotics for horses cannot be estimated, as such products are frequently sold on requisition and administered by the veterinarian in connection with a clinical examination in ambulatory practice, in clinics or in hospitals. 

### Dogs and cats

In 2025, the overall sales of veterinary medicinal products for oral medication of dogs were 502 kg, compared to 746 kg in 2016. The corresponding figures for cats were 93 and 73 kg, respectively. As in previous years, aminopenicillins (with and without clavulanic acid), first-generation cephalosporins and lincosamides were the classes with largest sales for dogs in 2025. For cats, products with aminopenicillins were by far the most sold (80%).  

The total number of packages dispensed for dogs in 2025, including both veterinary antibiotics and those authorised for use in humans, was <span class="nobr">125 366</span>, and for cats <span class="nobr">65 447</span>. The corresponding figures for 2006 were <span class="nobr">410 732</span> and <span class="nobr">140 067</span>, respectively. In 2006, sales for dogs corresponded to 563 packages per 1000 dogs. Since then, sales have decreased by 76% to 134 packages per 1000 dogs in 2025 (@fig-25). The most recent estimate of the number of dogs was conducted in January 2026. Additional estimates from 2012, 2017, 2020 and 2024 were also taken into account when interpolating population growth in the intervening years. Consequently, @fig-25 has been updated to reflect these revised figures. The most prominent changes relative to 2006 are noted for first generation cephalosporins (-90%), fluoroquinolones (-95%) and aminopenicillins with clavulanic acid (-86%). 

As described in [Svarm 2008](https://www.sva.se/djurhaelsa/antibiotika/oevervakning/swedres-svarm/){target="_blank"}, the emergence of infections with multiresistant methicillin-resistant *Staphylococcus pseudintermedius* and methicillin-resistant *S. aureus* triggered several national and local initiatives. This has most likely led to changes in prescribers’ behaviour, which in turn explains the downward trend in sales of antibiotics for dogs shown in @fig-25. 


The available estimates of numbers of cats are assumed to be more uncertain, hence no calculations to correct for population size were made.   

```{r}
#| include: false
# hidden chunk to read captions
dat_25 <- load_from_excel("data/Figure 2.5 Antibiotics for dogs.xlsx", sheet = "Data")
fig_25_cap <- dat_25$fig_cap
fig_25_alt <- dat_25$fig_alt 
```


::: {.panel-tabset}
###### Figure
```{r}
#| label: fig-25
#| echo: false
#| message: false
#| warning: false
#| fig-cap: !expr fig_25_cap
#| fig-alt: !expr fig_25_alt

names(dat_25$fig_data) <- c("Year", "Total")

df_fig_25 <- dat_25$fig_data |>
  mutate(tooltip = paste0("<b>Year:</b> ", Year, "<br><b>Packages per 1000 dogs:</b> ", Total), data_id = paste(Total, Year, sep = "_"),
          one_col = "Sales of antibiotics for oral medication")

yrs <- sort(unique(df_fig_25$Year))

#need to manually set the color to not have a black line
p_25 <- ggplot(df_fig_25, 
               aes(x = factor(Year),
                   y = Total,
                   group=one_col,
                   color=one_col)) +
  geom_line_interactive(aes(tooltip=glue::glue('{Year}'))) +
  ggiraph::geom_point_interactive(aes(tooltip = tooltip, data_id = data_id))+
  #scale_x_continuous(breaks = yrs, minor_breaks = NULL) +
  scale_y_continuous(limits = c(0, 650), breaks = seq(0, 600, by = 100))+
  #scale_color_SweSva() +
  scale_color_manual(values= one_color_SweSva[[1]]) +
  scale_y_continuous(limits = c(0,600))+
scale_x_discrete(expand=c(0,0.2)) +
  labs(x = "Year",
       y = "Packages per thousand dogs",) +
  theme_SweSva() +

  theme_SweSva_rotate() +

  add_sva_caption() 

make_dynamic_plot(plot = p_25, filename = "Fig2-5")
```

###### Data
```{r}
dat_25$fig_data |> 
  make_table_data_tab(file_name = "Fig2-5")
```
:::
