---
format:
  html:
    embed-resources: false
    toc: true

execute:
  cache: false
  echo: false
  warnings: false
  error: false

---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```

::: {.callout-note appearance="simple" collapse="false" #infoc-salestopical}
## [In-focus]{.in-focus} - Sales of topical antimicrobial products {.unnumbered}

### Sales of antimicrobial products formulated for topical use
  
Topical use of antimicrobial products in individual animals is expected to have a lower impact on antimicrobial resistance at population level than parenteral or oral use [@ema2025categorisation]. Therefore, topical products are not included in the scope of the monitoring of sales of antibiotics annually presented in Swedres-Svarm. Nevertheless, topical use of antimicrobials can select for resistance in individuals or (sub)populations and is part of antimicrobial stewardship. As trends in amounts and prescription patterns might differ compared to systemic treatments, a brief overview of sales of topical products with antibiotics, antifungals and antivirals for animals in Sweden is given. A comparison with historical data is also presented for products with at least one antibiotic substance. 

Data on sales of antimicrobial human and veterinary medicinal products were retrieved from the database of the Swedish eHealth Agency for the years 2020-2025. Only products sold for use in animals and formulated for topical use were included. For prescribed products, animal species as given on the prescription was included. “Antimicrobial” was defined as products with ATC codes defined as such by the [Medical Products Agency](https://www.lakemedelsverket.se/sv/lakemedel-for-djur/antimikrobiella-lakemedel){target="_blank"}. Compared to the inclusion as per Regulation (EU) 2021/578 [@eu2021reg578], products with only antiseptics, such as boric acid or chlorhexidine, were excluded, while products with antibiotics and/or antifungals with e.g. ATC codes (Q)D07C were included.  

The products sold were categorised as dermatological, ophthalmological or otological use based on ATC codes. Products with ATC codes in (Q)S03 were categorised as ophthalmological. The unit of measurement was number of packages sold for each product presentation. Product presentations formulated for long duration (seven-day dosing interval) were calculated to the smallest package size of the respective product.  

### Sales trend 2020-2025 

The number of packages of antimicrobial medicinal products sold for topical use in animals decreased by 19% between 2020 and 2025 (@fig-2IF1). The most marked decrease was for ophthalmologicals (-29%) and the least for otologicals (-15%). Throughout this period, ophthalmological products accounted for most of the sales. 

```{r}
#| include: false
# hidden chunk to read captions
dat_2IF1 <- load_from_excel("data/Fig1-focustopical-2020_2025.xlsx", sheet = "Underlag")
fig_2IF1_cap <- dat_2IF1$fig_cap
fig_2IF1_alt <- dat_2IF1$fig_alt 

```

::: panel-tabset

###### Figure
```{r}
#| label: fig-2IF1
#| message: false
#| warning: false
#| fig-cap: !expr fig_2IF1_cap
#| fig-alt: !expr fig_2IF1_alt


df_fig_413 <- dat_2IF1$fig_data |>
  pivot_longer(cols = c(Dermatological, Ophthalmological, Otological), names_to = "Category", values_to = "Value") 

dat_2IF1_long <- df_fig_413 |>
  mutate(Year = as.factor(Year), 
         tooltip = paste0("<b>Year:</b> ", Year, "<br><b>Category:</b> ", Category, "<br><b>Value:</b> ", round(Value, 1)), 
         data_id = paste(Year, Category, sep = "_"))

p_2IF1 <- ggplot(dat_2IF1_long, 
                 aes(x = Year,
                     y = Value, 
                     fill = Category,
                     tooltip = tooltip,
                     data_id = data_id)) +
  geom_bar_interactive(stat = "identity", position = "stack") +
  scale_fill_SweSva() +
  labs(x = "Year",
       y = "Number of packages", fill = NULL) +
  scale_y_continuous(limits = c(0, 120000), 
                     breaks = seq(0, 120000, by = 20000),
                     labels = scales::label_number(), expand = c(0,0.2)) +
  scale_x_discrete(expand = c(0,0.2)) +
  theme_SweSva() +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  add_sva_caption()
  
make_dynamic_plot(plot = p_2IF1, filename = "fig-2IF1")

```

###### Data
```{r}
df_fig_413 |>
  make_table_data_tab()
```
:::

### Descriptive analysis of data for 2025 

A more detailed breakdown for 2025 showed that the majority of the sold topical products were prescribed for dogs, followed by cats. Horses accounted for a markedly smaller proportion of sold packages (@fig-2IF2). The number of packages sold for production animals was negligible. In dogs, topical antimicrobials accounted for 30% of all dispensed prescriptions for antimicrobial products, compared with 17% in cats. 

Ophthalmological products accounted for the largest share of topical antimicrobial sales for dogs, cats and horses. A higher proportion of dermatological and otological products were sold for dogs compared to cats and horses (@fig-2IF2). 

```{r}
#| include: false
# hidden chunk to read captions
dat_2IF2 <- load_from_excel("data/Fig2-focustopical-species.xlsx", sheet = "data")
fig_2IF2_cap <- dat_2IF2$fig_cap
fig_2IF2_alt <- dat_2IF2$fig_alt 

```

::: panel-tabset

###### Figure
```{r}
#| label: fig-2IF2
#| message: false
#| warning: false
#| fig-cap: !expr fig_2IF2_cap
#| fig-alt: !expr fig_2IF2_alt

# no need to convert to long format as PowerBI already did this
dat_2IF2_long <- dat_2IF2$fig_data |>
    mutate(Species = factor(Species, levels = c("Dog", "Cat", "Horse", "Other", "Production animals")))

p_2IF2 <- ggplot(dat_2IF2_long, aes(x = Species, y = `Packages sold`, fill = `Target site`)) +
  geom_bar_interactive(aes(tooltip = paste0("<b>Species:</b> ", `Species`,
                                            "<br><b>Target site:</b> ", `Target site`,
                                            "<br><b>Value:</b> ", `Packages sold`),
                           data_id = paste(Species, `Target site`)), 
                       stat = "identity", position = "stack") +
  scale_fill_SweSva() +
  labs(x = "Species",
       y = "Number of packages", fill = NULL) +
  scale_y_continuous(limits = c(0, 50000), breaks = seq(0, 50000, by = 10000), 
                     labels = scales::label_number(),
                     expand = c(0,0.2)) +
    scale_x_discrete(expand = c(0,0.2)) +

  theme_SweSva() +
  theme(axis.text.x = element_text(hjust = 0.5)) +
  add_sva_caption()

  
make_dynamic_plot(plot = p_2IF2, filename = "fig-2IF2")
```

###### Data
```{r}
dat_2IF2_long |>
  make_table_data_tab()
```
:::

Sales of ophthalmologicals consisted almost exclusively of products with antibiotics. Sales of dermatologicals were dominated by products with antibiotics followed by antifungals (@fig-2IF3). Otologicals were dominated by products with antibiotic–antifungal combinations.  

Sales of topical antivirals were low across all sites of use. Of all the antiviral products sold, most (67%) were for horses, followed by cats (21%). Approximately two-thirds of products sold for horses were dermatological, while the remainder were ophthalmological. This pattern differed from that seen in cats, where only one-fifth of the products were dermatological and the remainder were ophthalmological. 


```{r}
#| include: false
# hidden chunk to read captions
dat_2IF3 <- load_from_excel("data/Fig3-focustopical-antimicrobial-effect.xlsx", sheet = "data")
fig_2IF3_cap <- dat_2IF3$fig_cap
fig_2IF3_alt <- dat_2IF3$fig_alt 

```

::: panel-tabset
###### Figure
```{r}
#| label: fig-2IF3
#| message: false
#| warning: false
#| fig-cap: !expr fig_2IF3_cap
#| fig-alt: !expr fig_2IF3_alt
# no need to convert to long format as PowerBI already did this
dat_2IF3_long <- dat_2IF3$fig_data
dat_2IF3_long <- dat_2IF3_long |>
  #workaround to determine the order in bars because scale_fill_SweSva overriders it
  mutate(stack_order = factor(`Type of antimicrobial effect`, levels = c("Antiviral", "Antifungal", "Antibiotic and antifungal", "Antibiotic")))

p_2IF3 <- ggplot(dat_2IF3_long, aes(x = `Target site`, y = `Packages sold`, fill = `Type of antimicrobial effect`, group = stack_order)) +
  geom_bar_interactive(aes(tooltip = paste0("<b>Target site:</b> ", `Target site`, 
                                            "<br><b>Type:</b> ", `Type of antimicrobial effect`,
                                            "<br><b>Value:</b> ", `Packages sold`),
                           data_id = paste(`Target site`, `Type of antimicrobial effect`)), 
                       stat = "identity", position = "fill") +
  scale_fill_SweSva() +
  labs(x = "Target site",
       y = "Number of packages (%)", 
       fill = NULL) +
  scale_y_continuous(labels = perc_100,#scales::percent_format(),
                     #limits = c(0,100),
                     expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0.2)) +
  theme_SweSva() +
  #theme(axis.text.x = element_text(hjust = 0.5)) +
  add_sva_caption()

  
make_dynamic_plot(plot = p_2IF3, filename = "fig-2IF3")
```

###### Data
```{r}
dat_2IF3_long |>
  make_table_data_tab()
```
:::

The majority (54%) of the topical antimicrobial products sold for animals in 2025 were human medicinal products prescribed for animals under the cascade [@eu2019reg6], while the remaining were veterinary medicinal products. Human medicinal products completely dominated the sales of ophthalmologicals, while most of the otologicals and dermatologicals were veterinary medicinal products (@fig-2IF3). The high proportion of human medicinal products for ophthalmologicals reflects the fact that, of the two veterinary medicinal products marketed in Sweden, one has not been available since early 2024 and the other not since October 2025.

```{r}
#| include: false
# hidden chunk to read captions
dat_2IF4 <- load_from_excel("data/Fig4-focustopical-VMP.xlsx", sheet = "data")
fig_2IF4_cap <- dat_2IF4$fig_cap
fig_2IF4_alt <- dat_2IF4$fig_alt 

```

::: panel-tabset

###### Figure
```{r}
#| label: fig-2IF4
#| message: false
#| warning: false
#| fig-cap: !expr fig_2IF4_cap
#| fig-alt: !expr fig_2IF4_alt

dat_2IF4_long <- dat_2IF4$fig_data |>
  #workaround to determine the order in bars because scale_fill_SweSva overriders it
  mutate(Target = factor(`Target site`, levels = c("Dermatological", "Ophthalmological", "Otological")), 
         stack_order = factor(`Product class`, levels = c("Veterinary medicinal product", "Human medicinal product")))

p_2IF4 <- ggplot(dat_2IF4_long, aes(x = `Target site`, y = `Packages sold`, fill = `Product class`, group = stack_order)) +
  geom_bar_interactive(aes(tooltip = paste0("<b>Target site:</b> ", `Target site`,
                                            "<br><b>Product class:</b> ", `Product class`,
                                            "<br><b>Value:</b> ", `Packages sold`),
                           data_id = paste(`Target site`, `Product class`)), 
                       stat = "identity", position = "stack") +
  scale_fill_SweSva() +
  labs(x = "Target site", y = "Number of packages", fill = NULL) +
  scale_y_continuous(limits = c(0, 50000), 
                     breaks = seq(0, 50000, by = 10000),
                     labels = scales::label_number(),
                     expand = c(0,0.2)) +
  scale_x_discrete(expand = c(0,0.2)) +
  theme_SweSva() +
  #theme(axis.text.x = element_text(hjust = 0.5)) +
  add_sva_caption()

make_dynamic_plot(plot = p_2IF4, filename = "fig-2IF4")
```

###### Data
```{r}
dat_2IF4_long |>
  make_table_data_tab()
```
:::



In 2025, most (79%) of the topical antimicrobial products were sold on prescription, while the rest (21%) were sold on requisition. When stratified by product category, ophthalmological products constituted the largest share of both prescription and requisition sales. Almost all products sold (98%) were authorised human or veterinary medicinal products, with the remainder consisting of products sold with a special license (2%).  

### Sales for dogs and cats - a comparison with historical data 

Data on pharmacy sales of products prescribed for topical use in dogs and cats from 2005-2007 were available from an earlier study [@Pettersson2017ex]. Only products containing at least one antibiotic substance were included. For comparison, corresponding data from 2023-2025 were extracted from the material above.  

The average yearly sales of dermatological and otological products with antibiotics for dogs and cats decreased by around 80% between the two periods (@fig-2IF5). For ophthalmological products, the decrease was less marked (38%), and the proportion of human medicinal products almost doubled. 

The marked decreases in sales of topical products with antibiotics for cats and dogs are in line with the declining trend of sales of antibiotic products for oral use (see Comments by species, Dogs and cats). National and local antimicrobial stewardship initiatives have most likely contributed to a change in practice, with emphasis on more precise diagnosis of underlying causes and the use of treatments without antibiotics when possible.  

```{r}
#| include: false
# hidden chunk to read captions
dat_2IF5 <- load_from_excel("data/Fig5-focustopical-historical.xlsx", sheet = "Data")
fig_2IF5_cap <- dat_2IF5$fig_cap
fig_2IF5_alt <- dat_2IF5$fig_alt 

```

::: panel-tabset
###### Figure
```{r}
#| label: fig-2IF5
#| message: false
#| warning: false
#| fig-cap: !expr fig_2IF5_cap
#| fig-alt: !expr fig_2IF5_alt

dat_2IF5_long  <- dat_2IF5$fig_data |>
  tidyr::pivot_longer(cols = c(`Veterinary medicinal products`,
                               `Human medicinal products`), 
                      names_to = "Product type", values_to = "Packages sold") |>
  mutate(`Animal species` = factor(`Animal species`, levels = c("Dogs", "Cats")),
         Period = factor(Period, levels = c("05-07", "23-25")),
         `Product type` = factor(`Product type`, levels = c("Veterinary medicinal products", "Human medicinal products")))



p_21F5 <- dat_2IF5_long |> 
  mutate(data_id = paste(Indication, `Animal species`, Period, `Product type`),
         tooltip_b = paste0("<b>Target site:</b> ", 
                            Indication, "<br><b>Species:</b> ", `Animal species`,
                            "<br><b>Period:</b> ", Period, 
                            "<br><b>Type:</b> ", `Product type`,
                            "<br><bValue:</b> ", round(`Packages sold`, 2))) |> 
  ggplot(aes(x = Period, 
             y = `Packages sold`, 
             fill = `Product type`,
             data_id = data_id)) +
  geom_bar_interactive(aes(tooltip = tooltip_b),
                       stat = "identity", position = "stack") +
  ggh4x::facet_nested(~ Indication + `Animal species`, scales = "free_x") +
  scale_fill_SweSva() +
  labs(x = "Years (2000)",
       y = "Number of packages (average)", 
       fill = NULL) +

  scale_y_continuous(limits = c(0, 80000), 
                     breaks = seq(0, 80000, by = 10000),
                     labels = scales::label_number(),
                     expand = c(0,0.2)) +
  theme_SweSva() +
   theme_SweSva_facet() +
  theme(plot.subtitle = element_text(vjust = -12),
        axis.text.x = element_text(angle=0,
                                   hjust = 0.5),
        strip.text = element_text(size=10)) +
  add_sva_caption() 

make_dynamic_plot(plot = p_21F5, filename = "fig-2IF5")
```


###### Data
```{r}
dat_2IF5_long |>
  make_table_data_tab()
```
:::

:::
