---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

<style>
.nobr {
  white-space: nowrap;
}
</style>


```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "polishing")
```

## Brief comments on data sources, methodology and confidentiality 

In Sweden, all veterinary medicinal products are sold by pharmacies. All pharmacies are obliged to report all sales of medicinal and veterinary medicinal products to the Swedish eHealth Agency, which maintains a database of sales from pharmacies to animal owners (prescriptions dispensed) and veterinarians (requisition for use in practice). 

For confidentiality reasons, sales of antibiotic classes with less than three products on the market have been aggregated in @tbl-21. 

The sales of veterinary medicinal products for mixing into feed used in aquaculture for food production are not included in the data referred to above, as such feed is traded from other countries. Data on prescriptions for fish are collected through a separate system, described in @sec-sales-fish Comments by animal species, Fish. 

Further details on data sources and inclusion criteria are given in @sec-mms - Materials and methods, sales of antibiotics.  

## Trends in animal populations 

Changes in the numbers of animals may affect trends in antibiotics sales statistics. Compared to 2016, the number of pigs slaughtered in 2025 was approximately the same, while the number of broilers increased by 13%. The number of dairy cows decreased by 11% during the same period. The number of horses was estimated to <span class="nobr">355 500</span> in 2016. In January 2026, the number of dogs was estimated to <span class="nobr">951 000</span> and the number of cats to <span class="nobr">1 582 000</span>. 

Further details on animal numbers and data sources can be found in @sec-demographics - Demographics and denominator data. 

## Completeness of data 

Until 2009, pharmacies in Sweden were run as a monopoly by a state-owned cooperation. In July 2009, the Swedish pharmacy market was re-regulated and many pharmacies compete on the market today. Some of these have niched in veterinary medicinal products and they are allowed to sell remotely to animal owners and veterinarians. 

At the time of the re-regulation, the responsibility of collecting sales data from pharmacies was transferred from the monopoly to a state-owned infrastructure company, and again in 2014 to the newly formed Swedish eHealth Agency. All pharmacies have a legal obligation to report all sales of medicinal and veterinary medicinal products to the Swedish eHealth Agency and are supervised by the Medical Products Agency. 

Between 2010 and 2015, there were two different problems resulting in a lack of completeness of data. Sales of products sold on special license were incomplete between 2011 and 2013 due to a system change. In 2013, concerns were also raised about a more general lack of completeness in the sales reported by pharmacies. The overall lack of completeness was estimated by SVA in collaboration with Marketing Authorisation Holders and was in the range of 5 to 10%. The problem persisted until 2015. A lack of completeness was also identified for 2017-2021. The cause was identified and corrected. Consequently, data for that period have been updated as of Swedres-Svarm 2022. Furthermore, the difference between 2021 and 2022 was inexplicably large (-12%). Despite a thorough search, no errors were identified, and there was no indication of a corresponding decrease in sales of non-antibiotic veterinary medicinal products. For more information, see [Swedres-Svarm 2022](https://www.sva.se/media/s3ggt1ny/swedres-svarm-2022-edit-230808.pdf){target="_blank"}.  

### Overall sales 

The total yearly sales of antibiotics for animals over the last decade are presented in @tbl-21. The potencies of different antibiotics are not equal and each class should therefore be evaluated separately.  

::: {.panel-tabset}
###### Table
```{r}
#| label: tbl-21
#| echo: false
#| message: false
#| warning: false

# Read source data from the Excel sheet
src_21 <- load_from_excel("data/Table 2.1 Sales AB 2025 total.xlsx", sheet = "Data")

tab_21 <- make_qt_table_from_excel("data/Table 2.1 Sales AB 2025 total.xlsx", sheet = "Data", footnotes = c("Includes small amounts of phenoxymethylpenicillin and penicillinase stable penicillins."))
tab_21$table

```

<!-- ###### julius's standard gt table -->

<!-- ```{r} -->
<!-- #| message: false -->
<!-- #| warning: false -->
<!-- data <- load_from_excel("data/Table 2.1 Sales AB 2025 total.xlsx", sheet = "Data") -->

<!-- df <- data$fig_data  -->
<!-- ``` -->


<!-- ```{r} -->
<!-- #| label: tbl_test1234 -->
<!-- #| tbl-cap: !expr glue::glue("{data$fig_cap}") -->
<!-- df |>  -->
<!--   gt()|>  -->
<!--   gt_theme_SweSva() -->
<!-- ``` -->




###### Data
```{r}
df <- src_21$fig_data
df |>
  make_table_data_tab(file_name = "tab21")
```
:::

In 2025, the proportion of sales on prescription, measured in kg active substance, was 63%, while the remainder was sold on requisition from pharmacies.

Of the overall sales expressed as kg active substance, around 90% are products formulated for treatment of individual animals (injectables, tablets, intramammaries etc.) and around 10% for treatment of groups or flocks (premixes, oral powders or solutions for in-water medication). In 2025, the total reported sales from Swedish pharmacies of antibiotics authorised for veterinary use were <span class="nobr">9 053</span> kg, of which 57% were benzylpenicillin. The corresponding figures for 2016 were <span class="nobr">10 165</span> kg and 55%, respectively. 

Since 2016, sales of most classes of antibiotics have decreased or remained relatively stable (@tbl-21). Compared to 2024, however, sales of aminopenicillins increased in 2025, mainly due to increased sales for poultry, as well as proportionally smaller increases in sales for pigs, dogs and cats. 

Since the withdrawal of veterinary medicinal products with high levels of zinc oxide in 2022, sales of aminoglycosides have increased in Sweden, reflecting an increased use of aminoglycosides for treatment of post-weaning diarrhoea in pigs. After reaching a peak in 2023, sales decreased in 2024 followed by a slight increase in 2025, see @sec-sales-pigs Comments on data by animal species, Pigs. 

#### Sales on special license 

The Swedish market for veterinary antibiotics is small, and for some therapy areas there are no suitable products authorised nationally. In such cases, products may be sold on special license and used in accordance with article 116 and articles 112-114 of Regulation (EU) 2019/6. An example is products for treatment of post-weaning diarrhoea in pigs via water, where aminoglycoside products from other countries are sold on special license. Furthermore, for some substance-formulation types, there are only one or two products with general marketing authorisation. When there are shortages of such products on the Swedish market, similar products sold on special license are used. For example, two nationally authorised injectable products with trimethoprim-sulphonamides were unavailable on the market from 2019 until late 2024. 

The proportion of sales of products on special license of the total sales of antibiotics for animals, in kg active substance, has fluctuated over the years. Between 2016 and 2025, that proportion varied between 3% and 14%, with the peak in 2023 and a subsequent decline to 8% in 2025 (@fig-21). The renewed availability of products with marketing authorisations in that class explains the reduction of products sold on special license in 2024 and 2025. 

```{r}
#| include: false
# hidden chunk to read captions
dat_21 <- load_from_excel("data/Figure 2.1 Sales on license.xlsx", sheet = "Data")
fig_21_cap <- dat_21$fig_cap
fig_21_alt <- dat_21$fig_alt 
```


::: {.panel-tabset}
###### Figure
```{r}
#| label: fig-21
#| echo: false
#| message: false
#| warning: false
#| fig-cap: !expr fig_21_cap
#| fig-alt: !expr fig_21_alt

names(dat_21$fig_data) <- c("Year", "Total")

df_fig_21 <- dat_21$fig_data |>
  mutate(tooltip_l = glue::glue('<b>{Year}</b>'),
         tooltip_p = paste0("<b>Year:</b> ", Year, "<br><b>Proportion (%) of LIC:</b> ", round(Total, 2)), data_id = paste(Total, Year, sep = "_"),
         one_col = "one_color_SweSva")

yrs <- sort(unique(df_fig_21$Year))

#need to manually set the line color to not have a black line

p_21 <- df_fig_21 |> 
  ggplot(aes(x = factor(Year), 
             y = Total, 
             group=one_col,
             color=one_col,
             data_id = data_id)) +
  geom_line_interactive(aes(tooltip=tooltip_l)) +
  geom_point_interactive(aes(tooltip = tooltip_p)) + #, color =  RColorBrewer::brewer.pal(9, "Set1")[3]) +
  scale_color_manual(values= one_color_SweSva) +
  scale_x_discrete(#breaks = yrs, minor_breaks = NULL,
    expand=c(0,0.2)) +
  scale_y_continuous(limits = c(0, 25),
                     breaks = perc_25,
                     expand=c(0,0),
                     #labels = function(x) paste0(x, "%")
                     ) +
  #scale_color_SweSva() +
  labs(x ="Year",
       y = "Proportion of total sales as kg (%)",
       color = NULL) +
  theme_SweSva() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
    legend.position = "none",
        plot.subtitle = element_text(vjust = +2)) +
  add_sva_caption()

make_dynamic_plot(plot = p_21, filename = "fig-21")
```

###### Data
```{r}
dat_21$fig_data |> 
  make_table_data_tab()
```
:::

### Population corrected data {#sec-sales-population}

To correct for changes in the numbers of animals over time, a population correction unit (PCU), described in a publication from the European Medicines Agency (EMA), was applied [@ema2011trends]. The PCU is a purely technical term representing an approximation of the summed live weight of the major animal populations, excluding companion animals. 

In @fig-22, the total sales of antibiotics for animals (including sales for companion animals) from 1980 onwards are presented as mg of active substance per PCU, using figures for 2024 as a proxy for PCU in 2025. As sales for use in aquaculture are not included in the data presented, fish have been excluded from the PCU given in the annual [ESUAvet reports](https://www.ema.europa.eu/en/veterinary-regulatory-overview/antimicrobial-resistance-veterinary-medicine/european-sales-use-antimicrobials-veterinary-medicine-esuavet-annual-surveillance-reports){target="_blank"} and the previous [ESVAC reports](https://www.ema.europa.eu/en/veterinary-regulatory-overview/antimicrobial-resistance-veterinary-medicine/european-surveillance-veterinary-antimicrobial-consumption-esvac-2009-2023){target="_blank"}. Another difference from data published in the ESUAvet and ESVAC reports is that data on products for use in companion animals are included in @fig-22. 


FIXME the footnotes need some kind of formatting solution.


```{r}
#| include: false
# hidden chunk to read captions
dat_22 <- load_from_excel("data/Figure 2.2 Sales for animals.xlsx", sheet = "Data")
fig_22_cap <- dat_22$fig_cap
fig_22_alt <- dat_22$fig_alt 
fig_22_fn <- c("Data from 2010-2015 are uncertain because of a lack of completeness mainly affecting injectable products. In the present figure, no attempts have been made to exclude products authorised for companion animals only, in contrast to data presented as mg per PCU in reports published by the European Medicines Agency.")
```

::: panel-tabset
###### Figure

```{r}
#| label: fig-22
#| message: false
#| warning: false
#| fig-cap: !expr make_fig_caption(fig_22_cap, fig_22_fn) # figure has a footnote so using a helper
#| fig-alt: !expr fig_22_alt

dat_22_long <- dat_22$fig_data |> 
  pivot_longer(cols = -Year, names_to = "Treatment type", values_to = "Value", values_drop_na = TRUE) |>
  mutate(`Treatment type` = factor(`Treatment type`, levels = c("Growth promotion", "Group treatment", "Individual treatment")))

df_fig_22 <- dat_22_long |>
  mutate(stack_order = dplyr::case_when(`Treatment type` == "Individual treatment" ~ 3,
                                        `Treatment type` == "Group treatment" ~ 2, 
                                        `Treatment type` == "Growth promotion" ~ 1), 
         tooltip = paste0("Year: ", Year, "<br>Treatment type: ", `Treatment type`, 
                          "<br>Value: ", round(Value, 2)), data_id = paste(`Treatment type`, Year, sep = "_"))

p_22 <- ggplot(df_fig_22,
               aes(x = factor(Year),
                   y = Value,
                   fill = `Treatment type`,
                   group = stack_order)) +
  geom_bar_interactive(aes(tooltip = tooltip, data_id = data_id), stat = "identity", position = "stack") +
  scale_fill_SweSva() +
  scale_y_continuous(expand = c(0,0.2), limits=c(0,50)) +
  labs(x = "Year",
       y = "mg per PCU", fill = NULL) +
  theme_SweSva() +
  theme_SweSva_rotate() +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  add_sva_caption()

make_dynamic_plot(plot = p_22, filename = "fig-22")

```

###### Data

```{r}
dat_22_long |>
  make_table_data_tab()
```
:::

Measured in mg per PCU, the overall sales were around 70% lower in 2025 compared to the average for 1980-1984 (i.e. before the Swedish ban on growth-promoting antimicrobials in 1986). This is explained first by the removal of growth-promoting antimicrobials in 1986, followed by a gradual but major decrease in the sales of veterinary products for group medication via feed or water from the mid-1990s onward. A decrease in sales of products for individual medication was also noted in the past two decades.

The Antimicrobial ad hoc expert group (AMEG) of EMA considers 3rd generation cephalosporins, fluoroquinolones and polymyxins to be classes of antibiotics for which there should be special restrictions regarding their use in animals (category B; restrict) [@ema2019categorisation]. Since 2016, the sales of these antibiotics, expressed as mg/PCU, have decreased considerably and remained below, or well below, 0.1 mg/PCU. For the 3rd generation cephalosporins and fluoroquinolones, the decrease is explained by a Swedish regulation that, since 2013, has limited veterinarians’ rights to prescribe or use these types of antimicrobials (SJVFS 2023:21). Regarding polymyxins, the findings of transferable resistance to colistin were communicated to stakeholders during 2016 and onwards. An awareness among prescribers of the importance of this class of antimicrobials for public health, and of the potential consequences of transferable resistance, is a probable explanation for the observed decrease. Use of colistin has increasingly been replaced with use of antibiotics in other classes, e.g. aminoglycosides. Since late 2023, veterinarians’ right to prescribe polymyxins has also been restricted by regulation SJVFS 2023:21. 
