---
format:
  html:
    embed-resources: false
    toc: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false

---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```

::: {.note .callout-infocus}

## [In-focus]{.in-focus} - MRSP clonal populations in Swedish dogs and cats {.unnumbered #infoc-mrsp}

### Methicillin-resistant *Staphylococcus pseudintermedius* (MRSP) clonal population structures in isolates from Swedish dogs and cats 


*Staphylococcus pseudintermedius* is part of the normal microbiota in dogs. It is associated with infections in a variety of body sites, but primarily with dermatologic infections such as pyoderma, otitis externa and wound infections, from which it is the most commonly isolated pathogen. Carriage and infections with *S. pseudintermedius* are also known to occur in cats, as are human colonisation and infection, although the latter are less commonly reported [@moses2023human]. 

The first methicillin-resistant *S. pseudintermedius* (MRSP) in Sweden was isolated in 2006. The numbers of isolated MRSP increased from twelve cases in 2006 to 130 cases in 2009. This called for changes in national recommendations aimed at reducing the prevalence of antibiotic-resistant bacteria in pets. These recommendations included enhanced infection prevention and control, continued routine sampling and bacterial culture with antibiotic susceptibility testing of most bacterial infections in dogs and cats, as well as policies aimed at reducing unnecessary use of antibiotics. Many Swedish veterinary hospitals and clinics quickly implemented these recommendations as well as subsequent regulations on infection prevention and control in veterinary care facilities and regulations on preventive measures aimed at reducing the risk of community-acquired MRSP carriage and infections. The numbers of MRSP cases decreased 2010-2011 to a relatively stable level of 40 to 60 cases each year. 

In Sweden, MRSP cases are notifiable and all index cases of MRSP isolated from animals must be sent to SVA and verified as MRSP by the presence of the *mecA* gene. In addition, all isolates are tested for antimicrobial susceptibility and typed phylogenetically using molecular typing methods. This is funded by the Swedish Board of Agriculture. By complementing already existing data regarding antibiotic susceptibility and genome sequencing, we presented changes over time, both in phenotypical antimicrobial resistance patterns and in the presence of antibiotic resistance genes, as well as presence of virulence genes [@myrenas2024genomic]. For the years 2012 to 2016, a randomized subset of 75% of the non-repetitive cases were included in the study. For the years 2017 to 2021, all reported non-repetitive cases were included. This resulted in a sample set of 356 isolates, of which 95% were of canine origin. Genotypical typing included staphylococcal chromosomal cassette *mec* (SCC*mec*) identification, multi-locus sequence typing (MLST) and whole genome MLST. 

The first detected MRSP in Sweden back in 2006 was later typed to ST71, a clone that was dominating in Europe at the time and also in Sweden until 2018. Over time, a considerable increase in the number of clonal lineages present was observed. The first Swedish *S. pseudintermedius* ST551 was isolated in 2015, became by 2018 the most common in Sweden and is still the most common in 2025. ST551 is on the rise also in the rest of Europe [@kizerwetter-swida2017changes; @papic2021genomic]. @fig-4IF1 shows the presence of the five most prevalent STs for the studied period and is an extension of the figure from the article published in 2024, also including the years 2022-2025. 

```{r}
#| include: false
# hidden chunk to read captions
dat_infocus_41 <- load_from_excel("data/Figure 1 In focus.xlsx", sheet = "data")
fig_infocus_41_cap <- dat_infocus_41$fig_cap
fig_infocus_41_alt <- dat_infocus_41$fig_alt 
```


::: {.panel-tabset}
###### Figure
```{r}
#| label: fig-4IF1
#| echo: false
#| message: false
#| warning: false
#| fig-cap: !expr fig_infocus_41_cap
#| fig-alt: !expr fig_infocus_41_alt

# keep same order as in xlsx file
st_order <- c("ST45", "ST71", "ST258", "ST265", "ST551", "Other STs")

# 100% stacked bar plot

dat_infocus_41_long <- dat_infocus_41$fig_data |>
  pivot_longer(cols = -Year, names_to = "ST", values_to = "value", values_drop_na = TRUE) |>
  mutate(Year = factor(Year), 
         ST = factor(ST, levels = st_order),
         value = as.numeric(value))


p_infocus_41 <- dat_infocus_41_long |>
  mutate(tooltip_b = paste0("<b>Year:</b> ", Year, "<br><b>Sequence type:</b> ", ST, "<br><b>Proportion:</b> ", value, "%"),
         data_id = paste(ST, Year, sep = "_")) |> 
  
  ggplot(aes(x = factor(Year),
             y = value,
             fill = ST,
             data_id = data_id,
             tooltip = tooltip_b)) +
  #ggiraph::geom_col_interactive(position = position_fill(reverse = TRUE)) +
        geom_bar_interactive(position="stack", stat="identity") +

  scale_y_continuous(#breaks = seq(0, 1, 0.1),
                     expand=c(0,0.1),
   # limits = c(0,100),
    #labels= scales::percent_format()
                     labels = function(x) x / 1
                     ) +
  scale_x_discrete(expand=c(0,0.2)) +
    guides(fill=guide_legend(nrow=1)) +

  scale_fill_SweSva() +
  labs(x = "Year", 
       y= "Proportion (%)", 
       fill = NULL) +
  theme_SweSva() +
  #theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  add_sva_caption()

make_dynamic_plot(plot = p_infocus_41, filename = "InFocus-MRSPanimals-Fig1")
```

###### Data
```{r}
dat_infocus_41_long |> 
  make_table_data_tab(file_name = "InFocus-MRSPanimals-Fig1")
```
:::

The change of dominant type over time is hard to explain but we found that ST71 isolates and ST551 isolates differ in a few ways. All the ST551 isolates, but only 12% of the ST71 isolates, carry the resistance genes *tet(K)* and/or *tet(M)*, giving resistance to doxycycline and tetracycline. In Sweden, tetracycline is not the primary treatment for MRSP infections [@sveriges2025] but it is used to some degree in neighbouring countries [@kizerwetter-swida2026tetracycline]. The ST551 isolates carry the gene coding for virulence factor SpsK while the ST71 isolates carry the gene for SpsL. These are surface proteins mediating binding to the host extracellular matrix but it is to our knowledge not known how this would affect the type prevalence. All isolates carried several virulence genes encoding factors associated with attachment, colonisation, toxin synthesis, quorum sensing, antibiotic resistance and immune evasion.

In total, 96% of the isolates from 2012-2021 were non-susceptible to, besides beta-lactams, at least one agent in three or more antibiotic categories. The percentages of isolates non-susceptible to erythromycin was 92%, clindamycin: 87%, trimethoprim-sulphamethoxazole: 86%, gentamicin: 72%, tetracycline: 67%, enrofloxacin: 62%. This is a cause of concern since it limits the treatment options. 

This In focus is a summary of [@myrenas2024genomic] with extended datasets and discussion. Further description of the study and a list of references can be found in the paper.

:::
