---
format:
  html:
    embed-resources: false
execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```


## Adverse reactions related to antibiotic use

Reported drug-related adverse reactions are continuously entered into BiSi, a national database administered by the Swedish Medical Products Agency. The reports originate both from healthcare professionals and patients. Adverse reactions related to antibiotics between 2021 and 2025 were analysed for various classes of agents.

There were 3 019 reports of side effects caused by the use of antibiotics during this period. The following organ system groups received most reports related to the use of systemic antibiotic drugs: skin and subcutaneous tissue disorders (n=1 258), gastrointestinal disorders (n=901), general disorders and administration site conditions (n=496), nervous system disorders (n=490), musculoskeletal and connective tissue disorders (n=208), respiratory, thoracic and mediastinal disorders (n=206), immune system disorders (n=143), infections and infestations (n=123), investigations (n=122), psychiatric disorders (n=119), reproductive system and breast disorders (n=119), renal and urinary disorders (n=110) and hepatobiliary disorders (n=102).

The majority of the reports during the five years period (67%) concern female patients, which corresponds to the gender difference seen in sales of antibiotics. The main antibiotic substances most commonly associated with adverse reactions in the last five years, unadjusted for sold substances and regardless of the cause of the report, are presented in @tbl-1-1.


::: panel-tabset
###### Table
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/tab-1-1.xlsx")
df <- data$fig_data |> 
  dplyr::select("Antibiotic":"Number of fatal cases")
```

```{r}
#| label: tbl-1-1
#| tbl-cap: !expr glue::glue("{data$fig_cap}")

df |> 
  gt() |> 
  gt_theme_SweSva()
```

###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Table1-1")
```
:::


