---
format:
  html:
    embed-resources: false
#    fig-cap-location: top

execute:
  cache: false
  echo: false
  warnings: false
  error: false
#bibliography: ../../references.bib
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "polishing")
```

After the exceptional changes in antibiotic sales during 2020 and 2021 – driven by the COVID-19 pandemic - total sales began to rise again in 2022 and continued to increase through 2023, while remaining lower than before the pandemic. The increase due to the rebound effect seen after the pandemic has slowed down as the sales remained substantially unchanged during 2024 and reduced in 2025 compared to the previous year. Throughout 2025, the epidemiological situation remained within normal parameters, with no significant deviations observed @FOHM2026.
 
The data sources and methodology underlying the statistics presented in this chapter are described in the Materials and methods, sales of antibiotics section. Due to regulations regarding the confidentiality of sales data, detailed data for substances and substance groups cannot be shown measured in DDD per 1 000 inhabitants. This affects the section for total sales as well as inpatient care.  


## Total sales of antibiotics


##### Results

- Total sales of antibiotics (J01 excl. methenamine) decreased by 2.8% compared to 2024 (from 10.97 DDD to 10.67 DDD per 1 000 inhabitants per day), @fig-1-1.
- Total sales of antibiotics varied between regions, ranging from 8.94 DDD per 1 000 inhabitants per day in Jönköping region to 11.78 DDD per 1 000 inhabitants per day in Blekinge region, @fig-1-1.
- Beta-lactamase sensitive penicillins (J01CE) and tetracyclines (J01AA) remain the two most sold antibiotic classes measured in DDD per 1 000 inhabitants in Sweden during 2025 (data not shown). The sales in outpatient care and inpatient care constitute 85% and 15% of the total sales, respectively. 


##### Comments

Nationally, the sales of antibiotics in 2025 remained below the sales volumes observed in 2019, prior to the COVID-19 pandemic. The sale levels in 2025 are lower compared to the previous year. The decrease was also observed regionally as sales volumes in 2025 decreased in 19 of 21 regions compared to 2024. The sales were higher in six of 21 regions compared to sales volumes observed in 2019. A comparison with the population-weighted mean of the EU/EEA countries from 2012-2022 @ECDC2023 and projectory over the following years confirms Sweden’s conservative position regarding antibiotic prescribing. Due to regulations regarding the confidentiality of sales data, detailed data for substance groups cannot be shown.



::: panel-tabset
```{r}
# Medskick: benchmarking för regionerna, de vill kunna jämföra mot varandra. Etablerad form att vi lägger dem i fallande ordning störst till minst på senaste året. 

data <- load_from_excel(path_file = "data/fig-1-1.xlsx")

df <- data$fig_data |> 
  dplyr::select(region,"2021":"2025") |> 
  pivot_longer(-region,names_to = "year") |> 
  mutate(sort_fct.tmp = case_when(region=="Sweden" ~ 1,
                              .default = 2))
```


###### Figure

```{r}
#| label: fig-1-1
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| message: false
#| warning: false

##| fig-cap: "Total sales of antibiotics (J01 excl. methenamine) in 2019-2023, by region."
spec_order <- df |> 
  filter(year==max(year)) |> 
  arrange(-value) |> pull(region)
  

p_ <- df |> 
  mutate(region = factor(region, levels = spec_order)) |> 
   mutate(data_id = row_number(),
          tooltip_b = glue::glue('<b>Year:</b> {year}
                                    <b>Region:</b> {region}
                                    <b>DDD/1 000 inhabitants per day:</b> {round(value,2)}')) |> 

  ggplot(aes(x = region,#fct_reorder(region,spec_order),
             y = value, 
             fill=factor(year),
             data_id = data_id)) +
    geom_col_interactive(aes(tooltip = tooltip_b),width = 0.5, position = position_dodge(0.7)) +

  #geom_bar_interactive(aes(tooltip = label_point,width = .9),
  #         position="dodge", stat="identity") +
  labs(x=NULL,
       y = "DDD/1 000 inhabitants per day") +
  scale_y_continuous(expand=c(0,.2),
                     labels = scales::label_number(),
                     limits = c(0,15)) +
  #scale_fill_fohm() +
  scale_fill_SweSva() + 
  theme_SweSva() +

  add_fohm_caption() +
  theme_SweSva_rotate() +
  scale_x_discrete(
    labels = ifelse(spec_order=="Sweden","**Sweden**",spec_order)
  ) +
  theme(axis.text.x = ggtext::element_markdown())
 

make_dynamic_plot(plot = p_, filename = "Figure-1-1")
```

###### Figure (alternative)

```{r}
#| warning: false
#| fig-height: 10
tmp_swe <-  df |> filter(region=="**Sweden**") |> transmute(year,
                                                        sweden=value)

p <-
  df |> 
  filter(region!="**Sweden**") |> 
  right_join(tmp_swe) |> 
  # rename(region = "value") |> 
  pivot_longer(cols=c(value,sweden)) |> 
  mutate(name_lab = case_when(name=="sweden"~"Sweden",
                              .default = region)) |> 
  mutate(data_id = row_number(),
         label_point = glue::glue('<b>{name_lab}</b>
                                   <b>{year}</b> 
                                    {round(value,2)} DDD/1 000 inhabitants per day')) |> 
  ggplot(aes(x=year,
             y=value,
             group=name,
             color=name,
             data_id = region
  )) +
  geom_line_interactive(aes(tooltip = glue::glue('<b>{name_lab}</b>'))) +
  geom_point_interactive(aes(tooltip = glue::glue('<b>{name_lab}</b>
                                                   <b>{year}</b> 
                                                  {round(value,2)} DDD/1 000 inhabitants per day'))) +
  scale_color_manual(values=fohm_diagram_discrete[1:2],labels = c("Sweden", "Region")) +
  facet_wrap(~ region, ncol=3) +
  
  labs(x = "Year",
       y = "DDD/1 000 inhabitants per day") +
  
  add_fohm_caption() +
  theme_SweSva() +
  theme_SweSva_facet() 

make_dynamic_plot(plot = p, filename = "test")
```


###### Data

```{r}
df |>
  make_table_data_tab(file_name = "test")
```
:::





