---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "polishing")
```

## Comparison of antibiotic resistance in human and veterinary medicine {#sec-cashv}


### ESBL-producing Enterobacterales

Enterobacterales with ESBL~A~ or ESBL~M~, and their corresponding genes, can transfer between animals and humans @efsa2011scientific, @debeen2014dissemination. The main route would be via food, but the possibility for direct transfer when handling animals should be kept in mind. 

The available data show that ESBL-producing bacteria are generally rare in animals and food in Sweden. Previously, the occurrence in intestinal samples from broilers was high but it has decreased considerably in recent years. Moreover, previous investigations when the occurrence was higher has shown that ESBL~A~- or ESBL~M~-producing *E. coli* constitute a small part of all the *E. coli* in the intestinal flora in a majority of the broiler samples. Finally, it has previously been shown that most isolates from humans in Sweden are not of the same types of ESBL~A~ or ESBL~M~ as in broilers. Hence, nothing indicates a need to revise the conclusion that food on the Swedish market is a limited source for ESBLs for humans @borjesson2016limited. Nevertheless, continued vigilance is warrented to prevent the development of reservoirs of ESBL-producing Enterobacterales in animals. 


### MRSA

Zoonotic transmission of MRSA occurs by direct or indirect contact. MRSA is reported globally in farm animals, companion animals, horses and wildlife. During the year, MRSA was isolated from pigs in a study with selective culture methods and sporadically in clinical samples from dogs, horses and cats. From dogs and cats, different *spa*-types were isolated, most of them previously found in humans. The situation among humans is also favourable.  

#### Livestock-associated MRSA 

In the last two decades, the zoonotic aspects on MRSA in farm animals has widened in many countries, largely due to spread of livestock-associated MRSA, and concerns pigs, veal calves, broilers and dairy cows. For pigs, the main clonal complex (CC) is 398. 

Based on surveillance of MRSA in livestock, with occasional findings in samples from cow, pig, goat and sheep, the situation has previously been considered favourable in Sweden. The prevalence in pigs is insufficiently studied, but in an EU baseline study with sampling of fattening pigs at slaughter and selective culture methods, MRSA was detected in 29% of investigated slaughter batches. In all positive batches, MRSA of CC398 was detected. *Spa*-type t034 was most common (n=50), followed by t011 (n=7) and a few t571 (n=3). How well this reflects the occurrence on farm is not known but sampling of farms with piglets is being performed in 2025/2026. Furthermore, MRSA CC398 occurs among horses and *spa*-type t011 (n=5) was the most common in 2025. 

In Sweden, human cases of MRSA CC398 are rare. In 2025, among all MRSA cases with available typing data, 23 cases were identified with the following *spa*-types: t034 (n=14), t011 (n=3), t1451 (n=2), t4652 (n=1), t4885 (n=1), t1456 (n=1), and t899 (n=1). Of these, ten isolates tested negative for PVL, while the PVL status of the remaining 13 isolates was unknown. Additionally, eleven isolates belonging to CC398 were detected through Regional whole genome sequencing. The possibility of animal contacts as a source is often not pursued. Consequently, epidemiological information regarding this is scarce.  

#### MRSA with *mec*C 

Isolates of MRSA with *mec*C were first reported internationally from dairy cows and humans in 2011 @garcia2011meticillin, @shore2011detection, @ito2012guidelines. Throughout the years, MRSA with *mec*C have been isolated from several animal species (cat, cow, dog, hedgehog, goat, pig and sheep). The total number of cases are low even if there are a number of isolates from hedgehogs in research projects and from goats in an outbreak at a zoo.  

Human cases of *mec*C-positive MRSA are also uncommon in Sweden. Five reported cases with the following *spa*-types were identified in 2025: t843 (n=2), t16761 (n=2) and t208 (n=1). In addition, another *mec*C case was typed to ST130 by whole genome sequencing. The epidemiological information concerning possible animal contacts is scarce but some of the *spa*-types in cases from humans have also been found in cases from animals. However, even if there would be zoonotic transfer, it is not currently considered a public health problem, as the number of cases of MRSA with *mec*C in humans in Sweden is low. 

#### MRSA-types typically associated with humans 

MRSA isolated from dogs and cats often belong to *spa*-types seen in MRSA from humans. This supports the view that humans often are the source of MRSA in companion animals @efsa2009scientific, @cvmp2009reflection. Spread can subsequently occur from animals to humans. However, the impact of companion animals as vectors for spread between humans is not known.  

#### Conclusions 

The MRSA situation in Sweden is in general still favourable both in humans and in animals. Biosecurity, with caution in trade of live animals and measures to prevent introduction by indirect routes, is important for preventing introduction and spread of MRSA in animal populations. However, with the detection of MRSA in the Swedish pig production, further emphasis need to be put not only on these matters to reduce further spread but also on hygiene measures to prevent spread from farms to the general public. Furthermore, antibiotic stewardship as well as hygiene and infection prevention and control measures are important to prevent healthcare-related spread between people, between animals or between people and animals. 

For more information on MRSA in Sweden, see @sec-arh (Antibiotic resistance in humans) and @sec-ara (Antbiotic resistance in animals). 

### MRSP 

*Staphylococcus pseudintermedius* may act as an opportunistic pathogen in humans and there are several reports in the literature of infections in humans with a varying degree of severity. However, MRSP is not generally considered to be a zoonotic pathogen. 

### VRE 

VRE have historically been isolated from a large proportion of broilers in Sweden using selective media. This occurrence has however decreased considerably. The occurrence in humans varies between years, mainly due to nosocomial outbreaks causing high occurrence in some years. However, based on genotypical investigations of isolates, there are no indications that the presence of VRE in broilers in Sweden has affected the situation in Swedish healthcare. 

### *Salmonella* 

Occurrence of *Salmonella* among farm animals, as well as among other animals, in Sweden is low and few incidents involve multiresistant strains. In 2025, the majority of isolates (92 of 108; 85%) were susceptible to all antibiotics tested. Resistance to fluoroquinolones (e.g. ciprofloxacin) is rare and in 2019, a strain with ESBL was detected for the first time, in an environmental sample from a farm. Thus, the overall situation in the veterinary sector is favourable, largely due to the strategies in the Swedish salmonella control programme initiated in the 1950s.  In 2025, half of the notifiable infections of *Salmonella enterica* were reported as domestic and 48% were acquired abroad. The origin of the isolates used in generating AST results from Svebar are not known. Considering the low occurrence of *Salmonella* in food-producing animals in Sweden, the majority of food-related infections presumably have a foreign source. The high occurrence of resistance to fluoroquinolones in isolates from humans (21%) in comparison to the very rare occurrence of such resistance in isolates from Swedish food-producing animals suggests that most of these isolates from human infections do not have a domestic origin. 


### *Campylobacter* 

Resistance to fluoroquinolones, tetracycline and erythromycin among faecal isolates of *Campylobacter jejuni* from humans was 56%, 30% and 0.8% respectively. From animals, 176 isolates of *C. coli* from healthy pigs were tested. The resistance found in *C. coli* from pigs against fluoroquinolones was 36% and one isolate was resistant to erythromycin. 

Resistance to erythromycin, the drug of choice for treatment of human campylobacteriosis, is rare among isolates from humans as well as animals in Sweden. In animals, macrolide resistance has only been found in five isolates, two from Swedish broiler meat in 2013 and one isolate each from pigs in 2017, 2023 and 2025. The isolates have been genome sequenced and no transferable macrolide resistance gene has been found. 

  

### Clinical resistance in  *Escherichia coli* from humans and animals

Comparison of resistance in bacteria from humans and different animal categories may indicate the magnitude of possible transfer of resistance between sectors and give insight into the drivers for resistance in the specific populations. However, in Swedres-Svarm, direct comparison of resistance is problematic as different interpretative criteria are used for bacteria from humans and animals. Data for bacteria from humans are interpreted with clinical breakpoints and presented as the proportion of isolates with clinical resistance. In contrast, data for bacteria from animals are mainly interpreted with epidemiological cut-off values (ECOFF) and presented as the proportion of isolates of non-wild type. For further information on interpretive criteria, see sections [Guidance for readers](#sec-g4r) and [Materials and methods](#sec-backdatamm). 

For the purpose of the comparison in this section, some data sets for *E. coli* from animals presented in Swedres-Svarm have been interpreted using clinical breakpoints for humans (@tbl-5-3). 

Resistance was generally more common in *E. coli* from humans than in isolates from animals (@tbl-5-3). Notably, clinical resistance to fluoroquinolones or 3rd generation cephalosporins was considerably more common in *E. coli* from humans than in isolates from animals, with the highest occurrence in blood stream isolates from humans (@tbl-5-3). This is in line with the very low use of these antibiotic classes in animals (see Sales of antibiotics for animals @sec-ara). However, although few isolates of *E. coli* from animals show clinical resistance to fluoroquinolones, reduced susceptibility (i.e. non-wild type) is more common in some categories of diseased and healthy animals (see Antibiotic resistance in animals @sec-asa) and previous reports). Possibly, the selection pressure from use of fluoroquinolones in animal populations is not sufficient to select for further mutations to clinical resistance in isolates with reduced susceptibility. 

For the antibiotics commonly used in both animals and humans, e.g. ampicillin and trimethoprim, resistance is more frequent. In particular, the occurrence of resistance is high among clinical isolates from calves, pigs and humans (@tbl-5-3, @fig-5-2 and @fig-5-3). When comparing resistance to trimethoprim, it should be considered that for some categories (i.e. clinical isolates from animals and blood isolates from humans), trimethoprim-sulphamethoxazole was tested. This could possibly result in a lower occurrence of resistance than if susceptibility to only trimethoprim had been tested.

Occurrence of resistance to ampicillin or trimethoprim could also be due to co-selection by use of other antibiotics or to other factors selecting for resistance. For example, although exact data are missing, use of ampicillin or amoxicillin in cattle is believed to be low in Sweden. Nevertheless, resistance to ampicillin is common in both isolates from diseased calves and dairy cows. However, it is well known that multiresistant *E. coli* is common in pre-weaned dairy calves but that resistant strains are cleared as calves mature. 

Moreover, the high occurrence of resistance to ampicillin or trimethoprim may be influenced in some categories by a low number of isolates and possible sampling bias, where animals are sampled due to therapeutic failures, inferring a selection of problematic cases. 

@fig-5-2 is Figure_5.2_Ampicillin.xlsx

::: panel-tabset
###### Figure
```{r}
data <- load_from_excel(path_file = "data/Figure_5.2_Ampicillin.xlsx")

df <- data$fig_data |>
  dplyr::select(1:3) |> 
  janitor::clean_names() |> 
  rename(value = ampicillin) |> 
  filter(!is.na(value))

```

```{r}
#| label: fig-5-2
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false
p <- df |> 
  mutate(#cluster = factor(cluster, levels=rev(spec_ordning)),
         data_id = row_number(),
         tooltip_b = glue::glue('<b>Category:</b> {category}
                                   <b>Proportion of clinical resistance:</b> {round(value,2)}%')) |> 
  
  ggplot(aes(x = value,
             y = fct_reorder(factor(category),value), 
             fill=factor(highlight),
             data_id = category)) +
  geom_col_interactive(aes(tooltip = tooltip_b),
                       width = 0.9,
                       position = position_dodge(0.7)) +
  labs(x = "Proportion of clinical resistance (%)",
       y = "Category") +
  scale_x_continuous(expand=c(0,.1), limits = c(0,100)) +
  scale_y_discrete(expand=c(0,0.1)) +
  #guides(fill=guide_legend(nrow=2)) +
  
  scale_fill_SweSva() + 
  theme_SweSva() +
  theme_SweSva_facet() +
  theme(axis.text.y = element_text(hjust=1)) +
  add_fohm_caption() +

   theme(axis.text.x = element_text(angle=0, vjust = 0, hjust = 0.5),
        panel.spacing.x = unit(1, "lines"),
        panel.grid.major.x = element_line(linetype = "solid", 
                                          linewidth = 0.25, 
                                          color = "#999999"),
        panel.grid.major.y = element_blank(),
        #plot.subtitle = element_text(vjust = +1)
        #,
        
    plot.margin = margin(0, .4, 0, 0, "cm")  # Top, Right, Bottom, Left
  )

make_dynamic_plot(plot = p, filename = "Fig5-2")
```
###### Data
```{r}
df |> 
  rename(Kingdom = highlight) |> 
  make_table_data_tab(file_name = "Fig5-2")
```
:::

@fig-5-3 is Figure_5.3_Trimethoprim.xlsx

::: panel-tabset
###### Figure
```{r}
data <- load_from_excel(path_file = "data/Figure_5.3_Trimethoprim.xlsx")

df <- data$fig_data |>
  dplyr::select(1:3) |> 
  janitor::clean_names() |> 
  rename(value = trimethoprim) |> 
  filter(!is.na(value))
```

```{r}
#| label: fig-5-3
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false
p <- df |> 
  mutate(#cluster = factor(cluster, levels=rev(spec_ordning)),
         data_id = row_number(),
         tooltip_b = glue::glue('<b>Category:</b> {category}
                                   <b>Proportion of clinical resistance:</b> {round(value,2)}%')) |> 
  
  ggplot(aes(x = value,
             y = fct_reorder(factor(category),value), 
             fill=factor(highlight),
             data_id = category)) +
  geom_col_interactive(aes(tooltip = tooltip_b),
                       width = 0.9,
                       position = position_dodge(0.7)) +
  labs(x = "Proportion of clinical resistance (%)",
       y = "Category") +
  scale_x_continuous(expand=c(0,.1), limits = c(0,100)) +
  scale_y_discrete(expand=c(0,0.1)) +
  #guides(fill=guide_legend(nrow=2)) +
  
  scale_fill_SweSva() + 
  theme_SweSva() +
  theme_SweSva_facet() +
  theme(axis.text.y = element_text(hjust=1)) +
  add_fohm_caption() +

   theme(axis.text.x = element_text(angle=0, vjust = 0, hjust = 0.5),
        panel.spacing.x = unit(1, "lines"),
        panel.grid.major.x = element_line(linetype = "solid", 
                                          linewidth = 0.25, 
                                          color = "#999999"),
        panel.grid.major.y = element_blank(),
        #plot.subtitle = element_text(vjust = +1),
        
    plot.margin = margin(0, .4, 0, 0, "cm")#,  # Top, Right, Bottom, Left
  #legend.position = "none"
  )

make_dynamic_plot(plot = p, filename = "Fig5-3")
```
###### Data
```{r}
df |> 
    rename(Kingdom = highlight) |> 
  make_table_data_tab(file_name = "Fig3-22")
```
:::

@tbl-5-3 is Table_5.3_resistance_various_Ecoli.xlsx

```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Table_5.3_resistance_various_Ecoli.xlsx")

df <- data$fig_data |> 
  head(n=14) |>
   mutate_if(is.numeric, round) |> 
  mutate("Tmp (>4)" = case_when(`Sample type`=="Urinary" & Year=="2025" ~ as.character(round(as.numeric(`Tmp (>4)`),2)),
                                .default = `Tmp (>4)`)) |> 
  mutate(across(everything(), as.character))


tbl_fn <- data$fig_data[16,1]
```

::: panel-tabset
###### Table
```{r}
#| label: tbl-5-3
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva(source_note = tbl_fn)
```

###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Table5-3")
```
:::



