---
format: html
---

```{r}
#| message: false
#| warning: false
source("../_common.R")
```

# Document structure  {.unnumbered}

# Headline 1 {.unnumbered}

Main text...

## Headline 2

Main text...

### Headline 3

Main text...

#### Headline 4

Main text...

##### Headline 5

Main text...

Or:

# Heading 1 - Document Title
## Heading 2 - Major Sections  
### Heading 3 - Subsections
#### Heading 4 - Minor Sections
##### Heading 5 - Rare Usage
###### Heading 6 - Very Rare


::: {.column-margin}
We know from *the first fundamental theorem of calculus* that for $x$ in $[a, b]$:

$$\frac{d}{dx}\left( \int_{a}^{x} f(u)\,du\right)=f(x).$$
:::

### Lists

#### Unordered Lists:

- First item
- Second item  
- Third item

#### Ordered Lists:

1. First step
2. Second step
3. Third step


[This is a span that has the class aside which places it in the margin without a footnote number.]{.aside}

::: {.panel-tabset}

##### Tab 1

Content of Tab 1

##### Tab 2

Content of Tab 2

##### Tab 3

Content of Tab 3

:::

You can include footnotes ^[This is a footnote.]

::: callout-note
## Highlight headline

One way to highlight something...
:::

::: {.callout-note}
## Note

Important information for readers.
:::

::: {.callout-warning}
## Highlight

Something to be careful about.
:::

or without title

::: {.callout-note .my-callout}
Some normal callout
:::


::: {.callout-tip}
Some wild callout
:::

::: {.callout-warning}
Some wild callout
:::


::: {.callout-note appearance="simple" collapse="true" color="red"}
## In-focus 


### Titel for in-focus

Main text...

:::


```{r}

p <- dummy_plot +
  labs(title = "FOHM figure") +
  scale_color_SweSva() + 
  add_fohm_caption() +
  theme_SweSva()

make_dynamic_plot(plot = p, filename = "FOHM_dummy")

```




::: panel-tabset
###### Figure
```{r}
#| label: fig-example0
#| fig-cap: !expr glue::glue("{}")
#| fig-alt: !expr glue::glue("{}")
#| warning: false
#| when-format: "html"

make_dynamic_plot(plot = p, filename = "FOHM_dummy")
```
###### Data
```{r}
mtcars |> 
  make_table_data_tab(file_name = "test")
```
:::


::: panel-tabset
###### Figure

###### Data
```{r}
mtcars |>   make_table_data_tab(file_name = "test")
```
:::



## Usefull functions

```{r}
source("../_common.R")
```

There are some useful functions included in `_common.R`.

-> loads all required libraries

- `load_from_excel()`
- `status()`

- `theme_SweSva()`
- `add_fohm_caption()`
- `add_sva_caption()`

- `make_dynamic_plot()`
- `make_table_data_tab()`
-

## Example usage:

```{r}
#| include: false
data <- load_from_excel("fig-exemple.xlsx")
```

## Create a plot

Required data structure:

```{r}
data$fig_data |> head()
```


```{r}
#| label: fig-example
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| echo: true

data$fig_data |> 
  pivot_longer(-region) |> 
  ggplot(aes(x=region, y=value, group=name, color=name)) +
  geom_point() +
  geom_line() +
  labs(y="y-axis label",
       x= "Region") +
  ## add standard theme, color/fill and caption
  theme_SweSva() +
    scale_y_continuous(breaks = scales::pretty_breaks()) +
  scale_color_SweSva() +
  add_fohm_caption() +
## additional modification, e.g. x-axis text 
    theme(axis.text.x = element_text(angle=90)) 
```


In the text, one can reference to the figure by placing `@fig-example` where you want the reference to be. This will result in the following cross-reference: @fig-example, that shows the figure when hovering over.

## Create a table with gt

Required data structure:

```{r}
data$fig_data |> head()
```

```{r}
#data$fig_data |>   make_table_data_tab(file_name = "test")
```


```{r}
#| label: tbl-example
#| tbl-cap: "Example"
#| echo: true

# Create the table with gt
test_insert <- "test"

data$fig_data |>
  gt() |> 
  cols_label(
    region = "Region",
    `2023` = html(glue::glue("<b>2023</b><br>(test = {test_insert})")),
    #urine = html(glue::glue("<b>Urine isolates, % R</b><br>(n= {n_urine})"))
    
  ) %>%
  tab_options(
    table.font.size = px(14),
    heading.title.font.size = px(16),
    heading.background.color = "lightgray"
  ) %>%
  fmt_markdown(columns = everything()) %>%
  # Left-align the first column (including label and rows)
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_column_labels(columns = region)
  ) %>%
  # Center-align the other columns (including labels and rows)
  tab_style(
    style = cell_text(weight = "bold",
                      align = "center"),
    locations = cells_column_labels(columns = c(`2020`,`2021`,`2022`,`2023`,`2024`))
  ) %>%
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_body(columns = c(`2020`,`2021`,`2022`,`2023`,`2024`))
  ) %>%
  opt_row_striping(row_striping = TRUE)
```


In the text, one can reference to the table by placing `@tbl-example` where you want the reference to be. This will result in the following cross-reference: @tbl-example, that shows the table when hovering over.


```{r}

# Apply the theme function with a source note
data$fig_data |>
  head(n=5) |> 
  gt() |> 
  gt_theme_fohm(
    source_note = "Statistiken uppdateras den 8:e varje månad med föregående månads inrapporterade fall."
  )

```




### Cross reference to other sections

```
## My Section Title {#sec-my-section}
```
In the text, add `@sec-my-section` to get something like this: @sec-material-methods.

## File download

[Download Image File](images/guidance_reader.png){download="guidance_reader.png"}

## download table

```{r}
tab_1 <- data$fig_data |>
  gt() |> 
  cols_label(
    region = "Region",
    `2023` = html(glue::glue("<b>2023</b><br>(test = {test_insert})")),
    #urine = html(glue::glue("<b>Urine isolates, % R</b><br>(n= {n_urine})"))
    
  ) %>%
  tab_options(
    table.font.size = px(14),
    heading.title.font.size = px(16),
    heading.background.color = "lightgray"
  ) %>%
  fmt_markdown(columns = everything()) %>%
  # Left-align the first column (including label and rows)
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_column_labels(columns = region)
  ) %>%
  # Center-align the other columns (including labels and rows)
  tab_style(
    style = cell_text(weight = "bold",
                      align = "center"),
    locations = cells_column_labels(columns = c(`2020`,`2021`,`2022`,`2023`,`2024`))
  ) %>%
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_body(columns = c(`2020`,`2021`,`2022`,`2023`,`2024`))
  ) %>%
  opt_row_striping(row_striping = TRUE)


#tab_1 |> gtsave(filename = "tab_1.html", inline_css = TRUE)
#[Download tab_1 File](tab_1.html){download="tab_1.html"}
```


## Flow diagrams in mermaid language

```{mermaid}
%%| label: quarto-mermaid-example-1
%%| fig-width: 5
flowchart LR
  qmd --> J([Jupyter])
  qmd --> K([knitr])
  J --> md
  K --> md
  md --> P([pandoc])
  P --> typst
  P --> html
  P --> docx
  
%% Make Node B link to Google
    click html href "https://en.wikipedia.org/wiki/HTML" "html wikipedia"
```

```{mermaid}
---
config:
  look: handDrawn
  theme: neutral
---
flowchart LR
  A[Start] --> B{Decision}
  B -->|Yes| C[Continue]
  B -->|No| D[Stop]
```

```{mermaid}
sankey

%% source,target,value
Electricity grid,Over generation / exports,104.453
Electricity grid,Heating and cooling - homes,113.726
Electricity grid,H2 conversion,27.14
```




::: {.callout-note appearance="simple" collapse="true" color="red"}
## In-focus DEMO

### This is a demo to illustrate in-focus


baskjdkjdskjhsdkjhskhjs^[This is a footnote-1.]

ruireiufkljfdkjfdjkfjhk^[This is a footnote-2.]

djljkdfjlkfdjkl^[This is a footnote-3.]

:::

### mardown styling in tooltip
###### Test
```{r}
# Vectorized Markdown → HTML conversion
md_to_html <- Vectorize(function(x) commonmark::markdown_html(x), USE.NAMES = FALSE)

p <- ggplot(mtcars, aes(
  x = mpg,
  y = wt,
  tooltip = md_to_html(paste("**MPG:**", mpg, "\n*Weight:*", wt))
)) +
  geom_point_interactive()
girafe(ggobj = p)
```
