---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```

## Demographics and denominator data {#sec-demographics}
### Humans

Official statistics on population in Sweden are provided by Statistics Sweden to the eHealth Agency. The number of people per age group and region in 2025 are provided in @tbl-6-1 and in total in Sweden from 2000-2025 in @tbl-6-2. These figures are used for calculation of antibiotic sales in humans.


```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Kap6_Demographics_table_6_1_2025_260511.xlsx")

df <- data$fig_data 
```

```{r}
#| label: tbl-6-1
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva() 
```

```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Kap6_Demographics_table_6_2_2025_260511.xlsx")

df <- data$fig_data 
```

```{r}
#| label: tbl-6-2
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva() 
```


### Animals

Official statistics on agriculture in Sweden are provided by the Board of Agriculture. The Board of Agriculture maintains a statistical database accessible online ([www.jordbruksverket.se](www.jordbruksverket.se){target="_blank"}). Annual figures on the number of animals are given in @tbl-6-3, on animals slaughtered in @tbl-6-4 and @tbl-6-5 and average herd size in @tbl-6-6. Readers are referred to the Board of Agriculture for further information. 

In brief, the number of dairy cows and pigs has decreased notably over the last three decades, but herd size has increased. During the same period, the number of beef cows has increased, as well as the number of chickens slaughtered. 

Estimates of the number of dogs and cats are available from the Board of Agriculture for 2006 and 2012, and in studies by the company Novus in 2017, 2020, 2024 and 2026. In 2012, the numbers of dogs and cats in Sweden were estimated to 784 000 and 1 159 000, respectively. The corresponding figures for 2026 were 951 000 and 1 582 000.   

 
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Kap6_Demographics_table_6_3_2025_260416.xlsx")

df <- data$fig_data |>
  mutate(across(-1, as.numeric),
         across(where(is.numeric), ~ round(., 0)))

```

```{r}
#| label: tbl-6-3
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva() 
```


```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Kap6_Demographics_table_6_4_2025_260416.xlsx")


df <- data$fig_data |> 
    mutate(across(-1, as.numeric),
         across(where(is.numeric), ~ round(., 0))) 
```

```{r}
#| label: tbl-6-4
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva() 
```

```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Kap6_Demographics_table_6_5 2025_260416.xlsx")

df <- data$fig_data |> 
     mutate(across(-1, as.numeric),
         across(where(is.numeric), ~ round(., 0))) 
```

```{r}
#| label: tbl-6-5
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva() 
```


```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Kap6_Demographics_table_6_6_2025_260416.xlsx")

df <- data$fig_data |> 
     mutate(across(-1, as.numeric),
         across(where(is.numeric), ~ round(., 0)))
```


```{r}
#| label: tbl-6-6
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva() 
```
