---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```

## *Acinetobacter* spp.

### *Acinetobacter* spp., from blood cultures

In 2025, 123 isolates of *Acinetobacter* spp. from blood were reported to Svebar. Carbapenem resistance was observed in 3.3% of the isolates (@fig-3-acino). Bloodstream infections caused by *Acinetobacter* spp. are still rare in Sweden compared to other countries in Europe, where multiresistant *Acinetobacter* spp. is a problematic pathogen in hospitals. Colistin resistance is occasionally seen in *Acinetobacter* and is mainly tested in multiresistant isolates, most of which have a link to health care abroad. 


### Microbiological surveillance programme, *Acinetobacter* spp.

In total, 37 unique (per patient, year, species and sequence type) isolates were received in 2025 within the microbiological surveillance programme for *Acinetobacter* spp. with reduced susceptibility to meropenem (I+R). This is a decrease compared to what was seen in 2024 (n=54), but similar to what was observed in 2023 (n=42). Of the 37 isolates, 28 carried species-specific *bla*~OXA~-genes (23, 24 or 58-like), eight harboured a combination of *bla*~OXA~ (23, 24 or 58-like) and *bla*~NDM~ and in one isolate, no carbapenemase gene was detected. Four clusters were active during 2025, with the number of unique isolates ranging from 2-6.




::: panel-tabset
###### Figure
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_Ny_Ace_spp_blood.xlsx")

df <- data$fig_data |>  
  mutate(across(everything(), as.character)) |>
    rename(antibiotic = Antibiotic) |> 

  pivot_longer(cols = -antibiotic, names_to = "year_tmp") |> 
  separate(year_tmp, into = c("year","feat"),sep = "-") |> 
  mutate(feat = str_remove(feat,"\r\n")) |> 
  mutate(value = as.numeric(value)) |> 
  pivot_wider(names_from = feat, values_from = value) |> 
    janitor::clean_names() |> 

## confidence interval
  rename(r_perc = percent_r,
         tot_antal = n_tested) |> 
  mutate(r_antal = round(tot_antal*(r_perc/100),0)) |> 
  group_by(antibiotic, year) %>%
  na.omit() |> 
  get_wilson_ci()
```

```{r}
#| label: fig-3-acino
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false

p <- df |>
  mutate(data_id=row_number(),
  tooltip_p = glue::glue('<b>Year:</b>{year}
                          <b>Antibiotic:</b>{antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}%
                          <b>Confidence interval:</b> {ci}'
                                ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= year,
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l)) +
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,linewidth = .5
  ) +
  scale_x_discrete(expand = c(0,0.2)) +
    scale_y_continuous(breaks = perc_25,
                       limits=c(0,25)) +
  labs(x="Year",
       y="Resistance (%)") +
  scale_color_SweSva() +
  theme_SweSva() +
  theme(plot.subtitle = element_text(vjust = -15)) +
  add_fohm_caption()

make_dynamic_plot(plot = p, 
                  filename = "Fig3-34")
```
###### Data
```{r}
df |> 
  transmute(antibiotic,
            year,
            "number of total tests" = tot_antal,
            "number of resistent tests" = r_antal,
            "% resistent" = r_perc,
            "Confidence interval" = ci) |> 
  make_table_data_tab(file_name = "Fig3-34")
```
:::


