---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```

## *Candidozyma auris*

### Microbiological surveillance programme, *Candidozyma auris* 


In January of 2025, a new microbiological characterisation programme for *Candidozyma auris* (formerly *Candida auris*) was introduced. *C. auris* has in recent years gained attention as a globally spread, multidrug-resistant yeast that causes severe healthcare-associated infections. 
In Sweden, *C. auris* has only occurred in sporadic cases among individuals who previously received hospital care abroad. In the previous 5 years, from 2020 to 2024, a total seven cases of *C. auris* have been reported in Sweden according to Svebar. 
Isolates from five patients, two women and three men age 20-77 years, were collected within the programme in 2025 and characterised with species confirmation and antifungal susceptibility results (AFST) at the national reference laboratory for fungi, at Karolinska University Hospital. The AFST results of these are shown in @tbl-3-cauris as resistance patterns. As expected, all isolates show elevated MICs for fluconazole, as the majority of *C. auris* strains have acquired resistance to the drug. Three isolates also show elevated MICs and resistance to anidulafungin, which is perhaps less expected, as *C. auris* resistance to echinocandins is less frequently reported (@tbl-3-cauris). 



```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Table_3.x_Cauris.xlsx",sheet = "data_ny")

df <- data$fig_data 
```

::: panel-tabset
###### Table
```{r}
#| label: tbl-3-cauris
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
   gt_theme_SweSva()
```

###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Table3-10")
```
:::

