---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "polishing")
```
## *Haemophilus influenzae*

### *Haemophilus influenzae*, from blood and nasopharynx cultures

-   Number of mandatory reported cases of invasive *H. influenzae*: 279

##### Comments

Invasive isolates of *H. influenzae* are notifiable according to the Communicable Disease Act regardless of antibiotic resistance. Cefotaxime resistance among invasive isolates remains low (@fig-3-30 and @tbl-3-hin). Among respiratory isolates, resistance levels to penicillin, ampicillin/amoxicillin, trimethoprim-sulfamethoxazole and cefotaxime increased during 2025 (@fig-3-28). 




@fig-3-hin-blood-ny is Figure_3._NY_Hin_blood

::: panel-tabset
```{r}
data <- load_from_excel(path_file = "data/Figure_3._NY_Hin_blood.xlsx")

df <- data$fig_data |> 
  #pivot_longer(-Antibiotic,names_to = "year")
   mutate(across(everything(), as.character)) |> 
  pivot_longer(cols = -Antibiotic, names_to = "year_tmp") |> 
  separate(year_tmp,into = c("year","feat"),sep = " - ") |> 
  mutate(value = as.numeric(value)) |> 
  pivot_wider(names_from = feat, values_from = value) |> 
janitor::clean_names() |> 
## confidence interval
  rename(r_perc =percent_r,
         tot_antal = n_tested) |> 
  mutate(r_antal = round(tot_antal*(r_perc/100),0)) |> 
  group_by(antibiotic, year) %>%
  na.omit() |> 
    get_wilson_ci()

```

###### Figure
```{r}
#| label: fig-3-hin-blood-ny
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false


p <- df |>
  mutate(data_id=row_number(),
   tooltip_p = glue::glue('<b>Year:</b> {year}
                          <b>Antibiotic:</b> {antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}%
                          <b>Confidence interval:</b> {ci}'
                                ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= year,
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l)) + #,tooltip = line_lab)) +
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,linewidth = .5
  ) +

  scale_x_discrete(expand = c(0,0.2)) +
  scale_y_continuous(breaks = perc_100,
                     limits=c(0,100)) +
  labs(x="Year",
       y="Resistance (%)") +
  scale_color_SweSva()+
  theme_SweSva() +
  add_fohm_caption() +
  guides(color=guide_legend(nrow=3)) +
  theme(plot.subtitle = element_text(vjust = -15))


make_dynamic_plot(plot = p, filename = "ny_hin_blood")
```
###### Data
```{r}
df |> 
  transmute(antibiotic,
            year,
            "number of total tests" = tot_antal,
            "number of resistent tests" = r_antal,
            "% resistent" = r_perc,
            "Confidence interval" = ci) |> 
  make_table_data_tab(file_name = "ny_hin_blood")
```
:::


@fig-3-28 is Figure_3.28_Hin_nph

::: panel-tabset
```{r}
data <- load_from_excel(path_file = "data/Figure_3.28_Hin_nph.xlsx")

df <- data$fig_data |> 
  #pivot_longer(-Antibiotic,names_to = "year")
  mutate(across(everything(), as.character)) |> 
  pivot_longer(cols = -Antibiotic, names_to = "year_tmp") |> 
  separate(year_tmp,into = c("year","feat"),sep = " - ") |> 
  mutate(value = as.numeric(value)) |> 
  pivot_wider(names_from = feat, values_from = value) |> 
  janitor::clean_names() |> 
  ## confidence interval
  rename(r_perc =percent_r,
         tot_antal = n_tested) |> 
  mutate(r_antal = round(tot_antal*(r_perc/100),0)) |> 
  group_by(antibiotic, year) %>%
  na.omit() |> 
    get_wilson_ci()

```

###### Figure
```{r}
#| label: fig-3-28
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false


p <- df |>
  mutate(antibiotic = gsub("\\(PcG 1\\)","",antibiotic)) |> 
  mutate(data_id=row_number(),
   tooltip_p = glue::glue('<b>Year:</b> {year}
                          <b>Antibiotic:</b> {antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}%
                          <b>Confidence interval:</b> {ci}'
                                ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= year,
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l)) + 
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,
                            linewidth = .5
                            ) +
  scale_x_discrete(expand = c(0,0.2)) +
    scale_y_continuous(breaks = perc_100,
                     limits=c(0,100)) +
  labs(x="Year",
       y="Resistance (%)") +
  scale_color_manual(values = col_SweSva_9) +
  theme_SweSva() +
  add_fohm_caption() +
  guides(color=guide_legend(nrow=3)) +
  theme(legend.text=element_text(size=10),
        plot.subtitle = element_text(vjust = -15))


make_dynamic_plot(plot = p, filename = "ny_hin_nph")
```
###### Data
```{r}
df |> 
  transmute(antibiotic,
            year,
            "number of total tests" = tot_antal,
            "number of resistent tests" = r_antal,
            "% resistent" = r_perc,
            "Confidence interval" = ci) |> 
  make_table_data_tab(file_name = "ny_hin_nph")
```
:::


@tbl-3-hin is Table_3.8_Hin

::: panel-tabset

```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Table_3.8_Hin.xlsx")

df <- data$fig_data |> 
  mutate(across(c(2:3), as.numeric))
```

###### Table
```{r}
#| label: tbl-3-hin
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
 gt_theme_SweSva()
```

###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Table3-8hin")
```
:::


