---
format:
  html:
    embed-resources: false
execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "polishing")
```

## Overview of sampling and culture results

Denominator data comes from Svebar. Over the last four years (2022–2025), we included data from 22 clinical laboratories, covering approximately 90% of Sweden’s population. @fig-3-2 shows annual counts for total cultures, blood cultures, urine cultures, nasopharyngeal cultures, and throat cultures. Reported culture numbers remained relatively stable throughout the period, with a slight decline in 2025 compared to 2024, partly due to one laboratory providing only 8 months of complete data.
The number of bacteria reported to EARS-Net yearly is shown in @fig-3-3.


::: panel-tabset
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.2_Svebar.xlsx") 
df <- data$fig_data |> 
  group_by(type,provtagnings_year) |> 
  summarise(n = sum(n)) |> 
  filter(provtagnings_year>2015)
```

###### Figure

```{r}
#| label: fig-3-2
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false

p <- df |> 
  
  mutate(data_id = row_number(),
         tooltip_b = glue::glue('<b>Sampling year:</b> {provtagnings_year}
                                    <b>Type:</b> {type}
                                    <b>Number of cultures:</b> {round(n,0)}')) |> 
  ggplot(aes(x=factor(provtagnings_year),
             y=n,
             group=type, 
             fill=type)) +
  geom_bar_interactive(aes(tooltip = tooltip_b),position="stack", stat="identity") +
  labs(y="Number of reported cultures",
       x="Year") +
  scale_y_continuous(expand=c(0,.2),
                     labels = scales::label_number()) +
  scale_x_discrete(expand = c(0,0.2)) +
  scale_fill_SweSva() + 
  theme_SweSva() +
  add_fohm_caption()

make_dynamic_plot(plot = p, filename = "Fig3-5")
```


###### Data
```{r}
df |> 
  rename(Samples_total = n) |> 
  make_table_data_tab(file_name = "Fig3-5")
```
:::


::: panel-tabset
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.3_EARS-Net.xlsx")
df <- data$fig_data |> 
  rename(feature = isolate) |> 
  pivot_longer(cols = -feature, names_to = "year")
```

###### Figure
```{r}
#| label: fig-3-3
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false
p <- df |> 
  mutate(data_id = row_number(),
         tooltip_b = md_to_html(glue::glue('<b>Year:</b> {year}
                                    <b>Species:</b> {feature}
                                    <b>Number of invasive isolates:</b> {round(value,0)}'))) |> 
  
  ggplot(aes(x = value,
             y = feature, 
             fill=as.factor(year),
             data_id = data_id)) +
  geom_col_interactive(aes(tooltip = tooltip_b),
                       width = 0.75, 
                       position = position_dodge(0.7)) +
  labs(y="Species",
       x = "Number of invasive isolates") +
  scale_x_continuous(expand = c(0,0.2),
  ) +
  scale_y_discrete(limits=rev) +
  scale_fill_SweSva() + 
  theme_SweSva(markdown = c("axis.text.y")) +
  add_fohm_caption() +
  theme(axis.text.x = element_text(angle=0, vjust = 0, hjust = 0.5),
        panel.spacing.x = unit(1, "lines"),
        panel.grid.major.x = element_line(linetype = "solid", linewidth = 0.25, 
                                          color = "#999999"),
        panel.grid.major.y = element_blank() ) +
  theme(plot.margin = margin(0, .4, 0, 0, "cm")  # Top, Right, Bottom, Left
  )

make_dynamic_plot(plot = p, filename = "Fig3-6")
```
###### Data
```{r}
df |> 
  rename("Notifiable resistance" = feature,
         "Number of cases" = value) |> 
  make_table_data_tab(file_name = "Fig3-6")
```
:::


