---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "polishing")
```

## Overview of sampling and culture results

Denominator data is derived from Svebar. For the last six years, denominator data from twelve clinical laboratories covering around 60% of the population in Sweden were included. In @fig-3-2, the annual numbers of analyses are presented for: total number of cultures (A), blood cultures (B), urine cultures (C), nasopharyngeal cultures (D) and throat cultures (E). Marginal decreases were noted compared to 2023 for number of cultures, blood cultures and urine cultures. Although there was a small increase in the number of nasopharyngeal cultures (3%) and throat cultures (10%), these are still fewer than in 2019. 

The number of bacteria reported to EARS-Net yearly is shown in @fig-3-3.


@fig-3-2 is Figure_3.2

::: panel-tabset
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.2_Svebar.xlsx") 
df <- data$fig_data |> 
  group_by(type,provtagnings_year) |> 
  summarise(n = sum(n)) |> 
  filter(provtagnings_year>2015)
```

###### Figure

```{r}
#| label: fig-3-2
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false

p <- df |> 
  
  mutate(data_id = row_number(),
         tooltip_b = glue::glue('<b>Sampling year:</b> {provtagnings_year}
                                    <b>Type:</b> {type}
                                    <b>Number of cultures:</b> {round(n,0)}')) |> 
  ggplot(aes(x=factor(provtagnings_year),
             y=n,
             group=type, 
             fill=type)) +
  geom_bar_interactive(aes(tooltip = tooltip_b),position="stack", stat="identity") +
  labs(y="Number of reported cultures",
       x="Year") +
  scale_y_continuous(expand=c(0,.2),
                     labels = scales::label_number()) +
  scale_x_discrete(expand = c(0,0.2)) +
  scale_fill_SweSva() + 
  theme_SweSva() +
  add_fohm_caption()

make_dynamic_plot(plot = p, filename = "Fig3-5")
```


###### Figure (alternative)

```{r}
p <- df |> 
  mutate(data_id = row_number(),
         tooltip_l = glue::glue('{type}'),
         tooltip_p = glue::glue('<b>Sampling year:</b> {provtagnings_year}
                                    <b>Type:</b> {type}
                                    <b>Number of cultures:</b> {round(n,0)}')) |> 
  
  ggplot(aes(x=provtagnings_year,
             y=n,
             group=type, 
             color=type)) +
  geom_point_interactive(aes(tooltip = tooltip_p)) +
  geom_line_interactive(aes(tooltip = tooltip_l)) +
  scale_x_continuous(breaks = c(2015:2025), 
                     expand = c(0,0.2)) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::label_number()) +
  scale_color_SweSva() +
  labs(y="Number of reported cultures",
       x="Year") +
  scale_fill_SweSva() + 
  theme_SweSva() 

make_dynamic_plot(plot = p, filename = "Fig3-5alt")
```

###### Figure per labb?

```{r}
#| height: 8

lab_order <- data$fig_data |> 
  group_by(labb_kod) |> 
  summarise(n = sum(n)) |> arrange(-n) |> distinct(labb_kod) |> pull(labb_kod)

p <- data$fig_data |> 
  mutate(labb_kod = factor(labb_kod, levels = lab_order),
         data_id = row_number(),
         tooltip_b = glue::glue('<b>Sampling year:</b> {provtagnings_year}
                                    <b>Type:</b> {type}
                                    <b>Number of cultures:</b> {round(n,0)}')) |> 
  ggplot(aes(x=factor(provtagnings_year),
             y=n,
             group=type, 
             fill=type)) +
  geom_bar_interactive(aes(tooltip = tooltip_b),position="stack", stat="identity") +
  labs(y="Number of reported cultures",
       x="Year") +
  facet_wrap(~labb_kod, ncol=7) +
  scale_x_discrete(expand = c(0,0.2)) +
  scale_y_continuous(expand=c(0,.2),
                     labels = scales::label_number()) +
  labs(y="Reported analysis (total)",
       x="Year") +
  
  scale_fill_SweSva() + 
  theme_SweSva() +
  theme_SweSva_facet() +
  add_fohm_caption()

make_dynamic_plot(plot = p, filename = "Fig3-5alt2")
```


###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Fig3-5")
```
:::




@fig-3-3 is Figure_3.3_EARS-Net

::: panel-tabset
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.3_EARS-Net.xlsx")
df <- data$fig_data |> 
  rename(feature = isolate) |> 
  pivot_longer(cols = -feature, names_to = "year")
```

###### Figure
```{r}
#| label: fig-3-3
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false
p <- df |> 
  mutate(data_id = row_number(),
         tooltip_b = md_to_html(glue::glue('<b>Year:</b> {year}
                                    <b>Species:</b> {feature}
                                    <b>Number of invasive isolates:</b> {round(value,0)}'))) |> 
  
  ggplot(aes(x = value,
             y = feature, 
             fill=as.factor(year),
             data_id = data_id)) +
  geom_col_interactive(aes(tooltip = tooltip_b),
                       width = 0.75, 
                       position = position_dodge(0.7)) +
  labs(y="Species",
       x = "Number of invasive isolates") +
  scale_x_continuous(expand = c(0,0.2),
  ) +
  scale_y_discrete(limits=rev) +
  scale_fill_SweSva() + 
  theme_SweSva(markdown = c("axis.text.y")) +
  add_fohm_caption() +
  theme(axis.text.x = element_text(angle=0, vjust = 0, hjust = 0.5),
        panel.spacing.x = unit(1, "lines"),
        panel.grid.major.x = element_line(linetype = "solid", linewidth = 0.25, 
                                          color = "#999999"),
        panel.grid.major.y = element_blank() ) +
  theme(plot.margin = margin(0, .4, 0, 0, "cm")  # Top, Right, Bottom, Left
  )

make_dynamic_plot(plot = p, filename = "Fig3-6")
```
###### Data
```{r}
df |> 
  rename("Notifiable resistance" = feature,
         "Number of cases" = value) |> 
  make_table_data_tab(file_name = "Fig3-6")
```
:::


