---
format:
  html:
    embed-resources: false
execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```


## *Pseudomonas aeruginosa*

Resistance levels are stable for most antibiotics in both blood isolates and non-respiratory isolates (@tbl-3-10, @fig-3-29 and @fig-3-30). Resistance to ceftazidime is most often due to efflux pumps and porin loss, not ESBL production. Tobramycin has replaced gentamicin as the recommended aminoglycoside. Colistin resistance is occasionally seen in *P. aeruginosa* and is mainly tested in multiresistant isolates, most of which have a link to health care abroad. 


### Microbiological surveillance programme, *Pseudomonas* spp. 

In total, 56 unique (per patient, year, species and sequence type) *Pseudomonas* spp. isolates were analysed in 2025, of which 15 belonged to the microbiological surveillance programme for ESBL~CARBA~-producing *Pseudomonas* spp. Eight of these isolates carried a *bla*~NDM~-gene, three *bla*~NDM~, two *bla*~VIM~ and two *bla*~IMP~. Three clusters were active during 2025, with four patients in the two largest clusters (*bla*~NDM~, *bla*~GES~) and two patients in the remaining cluster (*bla*~NDM~).


::: panel-tabset
###### Figure
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.29_Pae_blood.xlsx")

df <- data$fig_data |>  
  mutate(across(everything(), as.character)) |> 
  pivot_longer(cols = -`Antibiotic (ntest_%R)`, names_to = "year_tmp") |> 
  separate(year_tmp,into = c("year","feat"),sep = "_") |> 
  mutate(value = as.numeric(value)) |> 
  pivot_wider(names_from = feat, values_from = value) |> 
  rename(antibiotic = `Antibiotic (ntest_%R)`) |> 
  
## confidence interval
  rename(r_perc = R,
         tot_antal = ntest) |> 
  mutate(r_antal = round(tot_antal*(r_perc/100),0)) |> 
  group_by(antibiotic, year) %>%
  na.omit() |> 
    get_wilson_ci()
```

```{r}
#| label: fig-3-29
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false

p <- df |>
  mutate(data_id=row_number(),
 tooltip_p = glue::glue('<b>Year:</b>{year}
                          <b>Antibiotic:</b>{antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}% resistance
                          <b>Confidence interval:</b> {ci}'
                                ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= year,
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l)) + 
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,linewidth = .5
  ) +
  scale_x_discrete(expand = c(0,0.2)) +
    scale_y_continuous(breaks = perc_25,
                       limits=c(0,25),
                       expand=c(0,0.2)) +
  guides(color=guide_legend(nrow=3)) +
  labs(x="Year",
       y="Resistance (%)") +
  scale_color_SweSva() +
  theme_SweSva() +
      theme(plot.subtitle = element_text(vjust = -15)) +

  add_fohm_caption()

make_dynamic_plot(plot = p, 
                  filename = "Fig3-32")
```
###### Data
```{r}
df |> 
   transmute(antibiotic, year, "Total number" = tot_antal, "Resistance (%)"=r_perc, "95% CI" = ci) |> 
  make_table_data_tab(file_name = "Fig3-32")
```
:::


::: panel-tabset
##### Figure 
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.30_PAE_Othr_2016_2025.xlsx")

df <- data$fig_data |>  
  janitor::clean_names() |> 
  rename(r_perc = proportion_of_resistant_isolates,
         felstapel_low = felstapel_lag,
         felstapel_high = felstapel_hog,
         tot_antal=number_of_isolats_with_ast) |> 
  mutate(ci = paste0(x95_percent_lag,"-",x95_percent_hog),
         r_perc = r_perc*100)
```

```{r}
#| label: fig-3-30
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false
p <- df |>
  mutate(data_id=row_number(),
         tooltip_p = glue::glue('<b>Year:</b>{year}
                          <b>Antibiotic:</b>{antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}% resistance
                          <b>Confidence interval:</b> {ci}'
                                ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= factor(year),
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l)) +
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,linewidth = .5
  ) +
  scale_x_discrete(expand = c(0,0.2)) +
    scale_y_continuous(breaks = perc_25,
                       limits=c(0,25)) +

  labs(x="Year",
       y="Resistance (%)") +
  scale_color_SweSva() +
  theme_SweSva() +
  theme(plot.subtitle = element_text(vjust = -15)) +
  add_fohm_caption()

make_dynamic_plot(plot = p, 
                  filename = "Fig3-33")
```

##### Data
```{r}
df |> 
    transmute(antibiotic, year, "Total number" = tot_antal, "Resistance (%)"=r_perc, "95% CI" = ci) |> 
   make_table_data_tab(file_name = "Fig3-33")
```
:::




```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Table_3.10_Pae.xlsx")

df <- data$fig_data |> 
  mutate(across(c(2:3), as.numeric))
```

::: panel-tabset

###### Table
```{r}
#| label: tbl-3-10
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva() 
```

###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Table3-9")
```
:::


