---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "polishing")
```

## *Streptococcus agalactiae*

### *Streptococcus agalactiae*, from blood cultures

#### Comments

*S. agalactiae* is not included in the Communicable Disease Act. It is an important pathogen in the context of pregnancy and childbirth and can cause serious infections among others as well, mainly elderly with predisposing disease. Resistance to erythromycin and clindamycin is now 18% and 16%, respectively (@fig-3-32).

@fig-3-32 is Figure_3.32_25

::: panel-tabset
##### Figure 
```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.32_25.xlsx")

df <- data$fig_data |>  
  rename(antibiotic = Antibiotic) |> 
   dplyr::select(1:`2025 - % R`) |> 
  head(n=3) |> 

  pivot_longer(cols = -antibiotic, names_to = "year_tmp") |> 
  separate(year_tmp, into = c("year","feat"),sep = "-") |> 
  mutate(feat = str_remove(feat,"\r\n")) |> 
  mutate(value = as.numeric(value)) |> 
  pivot_wider(names_from = feat, values_from = value) |> 
    janitor::clean_names() |> 

## confidence interval
  rename(r_perc = percent_r,
         tot_antal = n_tested) |> 
  mutate(r_antal = round(tot_antal*(r_perc/100),0)) |> 
  group_by(antibiotic, year) %>%
  na.omit() |> 
  get_wilson_ci()
```

```{r}
#| label: fig-3-32
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false
p <- df |>
  mutate(data_id=row_number(),
tooltip_p = glue::glue('<b>Year:</b> {year}
                          <b>Antibiotic:</b> {antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}%
                          <b>Confidence interval:</b> {ci}'
                                ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= year,
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l),
                        lwd=0.5) +
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,
                            linewidth = .5
  ) +
  scale_x_discrete(expand = c(0,0.2)) +
  scale_y_continuous(breaks = perc_25,
                     limits=c(0,25)) + #c(-1,1)) + #scales::pretty_breaks()) +

  labs(x="Year",
       y="Resistance (%)") +
  scale_color_SweSva() +
  theme_SweSva() +
  #theme(plot.subtitle = element_text(vjust = -15)) +
  add_fohm_caption()

make_dynamic_plot(plot = p, 
                  filename = "Fig3-32")
```
##### Data
```{r}
df |> 
   make_table_data_tab(file_name = "Fig3-32")
```
:::

