---
format:
  html:
    embed-resources: false

execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```

## *Streptococcus pyogenes*

### *Streptococcus pyogenes*, from blood cultures

##### Results

-	Number of mandatory reported cases of invasive *S. pyogenes*: 750

##### Comments

Invasive cases of *S. pyogenes* are notifiable according to the Communicable Disease Act. In 2025, the incidence fell to 7.1 (cases/100 000 inhabitants), down from 12.9 in 2024 (n= 1 367). AST data for 509 isolates were provided by Svebar (@fig-3-31). Since 2023, resistance to clindamycin and erythromycin has stabilised at prepandemic levels and are now 4.5% and 6.3%, respectively. Some laboratories did not test susceptibility to trimethoprim-sulphamethoxazole and tetracycline. The variation in resistance during 2021 to 2022 should be viewed cautiously due to limited number of isolates. 


```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.31_25.xlsx")

df <- data$fig_data |>  
  mutate(across(everything(), as.character)) |>
    rename(antibiotic = Antibiotic) |> 
  dplyr::select(1:`2025 - % R`) |> 
  head(n=5) |> 

  pivot_longer(cols = -antibiotic, names_to = "year_tmp") |> 
  separate(year_tmp, into = c("year","feat"),sep = "-") |> 
  mutate(feat = str_remove(feat,"\r\n")) |> 
  mutate(value = as.numeric(value)) |> 
  pivot_wider(names_from = feat, values_from = value) |> 
    janitor::clean_names() |> 

## confidence interval
  rename(r_perc = percent_r,
         tot_antal = n_tested) |> 
  mutate(r_antal = round(tot_antal*(r_perc/100),0)) |> 
  group_by(antibiotic, year) %>%
  na.omit() |> 
  get_wilson_ci()
```
::: panel-tabset
###### Figure
```{r}
#| label: fig-3-31
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false

p <- df |>
  mutate(data_id=row_number(),
          tooltip_p = glue::glue('<b>Year:</b> {year}
                          <b>Antibiotic:</b> {antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}%
                          <b>Confidence interval:</b> {ci}'
                                ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= year,
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l),
                        lwd=0.5) +
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,linewidth = .5
  ) +
  scale_x_discrete(expand = c(0,0.2)) +
    scale_y_continuous(breaks = perc_100,
                       limits=c(0,50)) +
  guides(color=guide_legend(nrow=2)) +

  labs(x="Year",
       y="Resistance (%)") +
  scale_color_SweSva() +
  theme_SweSva() +
    theme(plot.subtitle = element_text(vjust = -15)) +

  add_fohm_caption()


make_dynamic_plot(plot = p, 
                  filename = "Fig3-35")
```
###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Fig3-35")
```
:::

