---
format:
  html:
    embed-resources: false
execute:
  cache: false
  echo: false
  warnings: false
  error: false
---

```{r}
#| results: "asis"
#| echo: false
#| warning: false
#| message: false
source("../../_common.R")
status(type = "ready2publish")
```

## *Salmonella*

### Mandatory reporting of *Salmonella*

Infection with Salmonella species are divided into three notifiable diseases in Sweden: infection with *Salmonella enterica* (*S*. Typhi and *S*. Paratyphi excluded), typhoid fever and paratyphoid fever. In addition, cases with Salmonella carrying ESBL or ESBL~CARBA~ are also notifiable in the mandatory reporting of ESBL-producing Enterobacterales. 

#### Results

- Total number of reported cases with *Salmonella enterica*: 1 431 (previous year: 1 612)
-	Total number of reported cases with typhoid fever: 16 (previous year: 26)
-	Total number of reported cases with paratyphoid fever: 24 (previous year: 9)
-	Total number of Salmonella carrying ESBL: 16 (previous year: 15)
-	Total number of Salmonella carrying ESBL~CARBA~: 0 (previous year: 1)

#### Comments

In 2025, half of the notifiable Salmonella infections were acquired in Sweden and almost half (48%) of the cases acquired salmonellosis abroad. For 2%, information was lacking. There have been five outbreaks with ten or more cases during the year, one of which included more than 100 cases with *S*. Enteritidis. This outbreak was linked to domestically produced eggs. The second largest outbreak had 48 cases infected with either of four different serotypes (*S*. Typhimurium, *S*. Richmond, *S*. Kinondoni, *S*. Newport) included. This outbreak was international and linked to alfalfa sprouts. No case was reported with Salmonella carrying ESBL~CARBA~.


### *Salmonella*, from blood or faecal and urine samples 

A total of 1 790 *Salmonella enterica* isolates were reported in Svebar, with 73% (n= 1 306) from faecal samples, 18% (n= 318) from blood and 6% (n= 104) from urine. For these sampling materials, approximately half had AST results reported. A comparison for 2025 is presented in @tbl-3-15.

Since 2021, the number of isolates found in blood with AST results has ranged between 45-152 per year and antibiotic (@fig-3-salmo1). The data may contain duplicates and there is a risk of overestimation of the resistance. Hence, results should be interpreted with caution. The general increase in resistance among faecal and urine isolates seen in 2022 (@fig-3-37) was probably linked to the increase of Salmonella isolates carrying ESBL. This increase then declined in the following years. In 2025, no significant changes were seen in resistance. Almost four-fifths of the Salmonella from faecal and urine samples were fully susceptible to azithromycin, cefotaxime and ciprofloxacin (@tbl-3-17). 



```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Figure_3.37_SalmonellaSpp_FecesUrin_2016_2025.xlsx")

df <- data$fig_data |>
  janitor::clean_names() |> 
  select(-c(no_of_laboratories,identifier)) |> 
  mutate(antibiotic = str_to_title(str_to_lower(antibiotic))) |> 
  ## confidence interval
  rename(r_perc = salm_spp_percent_r,
         tot_antal = number_of_isolats_with_ast) |> 
  mutate(r_antal = round(tot_antal*(r_perc/100),0)) |> 
  group_by(antibiotic, year) %>%
  na.omit() |> 
  get_wilson_ci()
```

::: panel-tabset
###### Figure
```{r}
#| label: fig-3-37
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| warning: false

p <- df |>
  mutate(data_id=row_number(),
         tooltip_p = glue::glue('<b>Year:</b> {year}
                          <b>Antibiotic:</b> {antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}%
                          <b>Confidence interval:</b> {ci}'
         ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= factor(year),
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l),
                        lwd=0.5) +
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,linewidth = .5
  ) +
  scale_x_discrete(expand = c(0,0.2)) +
  scale_y_continuous(breaks = perc_100,
                     limits = c(0,50)
  ) +
  guides(color=guide_legend(nrow=3)) +
  
  labs(x="Year",
       y="Resistance (%)") +
  scale_color_SweSva() +
  theme_SweSva() +
  theme(plot.subtitle = element_text(vjust = -15),
        legend.text = element_text(size=10)) +
  
  add_fohm_caption()

make_dynamic_plot(plot = p, 
                  filename = "Fig3-44")
```
###### Data
```{r}
df |> 
   transmute(antibiotic, year, "Total number" = tot_antal, "Resistance (%)"=r_perc, "95% CI" = ci) |> 
  make_table_data_tab(file_name = "Fig3-44")
```
:::




```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Table_3.15_Salmonella_Blod_FU_jmf2025.xlsx")

df <- data$fig_data |> 
  mutate(across(c(2:3), as.numeric))
```

::: panel-tabset
###### Table
```{r}
#| label: tbl-3-15
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva()
```

###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Table3-11")
```
:::



```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Table_3.16_Salmonella_blod_21_25.xlsx")

df <- data$fig_data |>
  janitor::clean_names() |> 
  ## confidence interval
  mutate(r_perc = as.numeric(percent_r),
         r_antal = number_of_resistant_isolates,
         tot_antal = n_tested) |> 
  mutate(r_antal = round(tot_antal*(r_perc/100),0)) |> 
  group_by(antibiotic, year) %>%
  na.omit() |> 
  get_wilson_ci()

```

::: panel-tabset
###### Figure 
```{r}
#| label: fig-3-salmo1
#| fig-cap: !expr glue::glue("{data$fig_cap}")
#| fig-alt: !expr glue::glue("{data$fig_alt}")
#| message: false
#| warning: false

p <- df |>
  mutate(data_id=row_number(),
         tooltip_p = glue::glue('<b>Year:</b> {year}
                          <b>Antibiotic:</b> {antibiotic}
                          <b>Resistance:</b> {round(r_perc,2)}%
                          <b>Confidence interval:</b> {ci}'
         ),
         tooltip_l = glue::glue('{antibiotic}')) |> 
  ggplot(aes(x= factor(year),
             y=r_perc,
             color = antibiotic,
             data_id = antibiotic)
  ) +
  geom_line_interactive(aes(group=antibiotic, tooltip=tooltip_l),
                        lwd=0.5) +
  geom_point_interactive(aes(tooltip = tooltip_p),hover_nearest = TRUE) +
  geom_errorbar_interactive(aes(ymin = r_perc - felstapel_low,
                                ymax = r_perc + felstapel_high,
                                tooltip = tooltip_p), 
                            width=.15,linewidth = .5
  ) +
  scale_x_discrete(expand = c(0,0.2)) +
  scale_y_continuous(breaks = perc_100, 
                     limits = c(0,50)
  ) +
  guides(color=guide_legend(nrow=3)) +
  
  labs(x="Year",
       y="Resistance (%)") +
  scale_color_SweSva() +
  theme_SweSva() +
  theme(plot.subtitle = element_text(vjust = -15),
        legend.text = element_text(size=10)) +
  add_fohm_caption()

make_dynamic_plot(plot = p, 
                  filename = "Fig3-45")
```

###### Data
```{r}
df |> 
  transmute(antibiotic, year, "Total number" = tot_antal, "Resistance (%)"=r_perc, "95% CI" = ci) |> 
  make_table_data_tab(file_name = "Fig3-45")
```
:::


```{r}
#| message: false
#| warning: false
data <- load_from_excel(path_file = "data/Table_3.17_Salmonella_kombres_21_25.xlsx")
df <- data$fig_data 
```

::: panel-tabset
###### Table
```{r}
#| label: tbl-3-17
#| tbl-cap: !expr glue::glue("{data$fig_cap}")
df |> 
  gt()|> 
  gt_theme_SweSva()
```

###### Data
```{r}
df |> 
  make_table_data_tab(file_name = "Table3-12")
```
:::

